---
title: "CAF subtype analysis"
author: "Kevin Ryan"
date: "`r Sys.time()`"
bibliography: citations.bib
output: 
  github_document:
     toc: true
     toc_depth: 3 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Cancer-associated fibroblasts (CAFs) are a heterogeneous cell type found in the tumour microenvironment. They have a wide array of functions, and tend to be immunosuppressive and cancer-promoting. There have been many attempts to characterise subtypes of CAFs, with much transcriptomic analysis being carried out in the Mechta-Grigoriou lab in Institut Curie. They have identified 4 'subtypes' which can be separated based on the expression of different markers:

 -  S1: FAP^High^, CD29^Med-High^, α^SMAHigh^, PDPN^High^, PDGFRβ^High^
 -  S2: FAP^Neg^, CD29^Low^, αSMANeg-^Low^, PDPN^Low^, PDGFRβ^Low^
 -  S3: FAP^Neg-Low^, CD29^Med^, αSMA^Neg-Low^, PDPN^Low^, PDGFRβ^Low-Med^
 -  S4: FAP^Low-Med^, CD29^High^, αSMA^High^, PDPN^Low^, PDGFRβ^Med^

[@Pelon2020]

FACS gating strategies can be used to isolate these various subtypes. The Mechta-Grigoriou group have done this and have generated bulk RNA-sequencing data for the S1, S3 and S4 subtypes. They generated scRNA-sequencing data for the S1 subtype. This data was deposited on the European Genome Phenome Archive, and was accessed via a Data Transfer Agreement.

The following summarises the data obtained:

+---------------+---------------+-----------------------+-------------------------------+
| Subpopulation | Total samples | Studies (Samples)     |  Notes                        |
+===============+===============+=======================+===============================+
| S1            |  28           | - EGAD00001003808 (16)| -  3808 has 12xJuxta-tumor    |
|               |               | - EGAD00001005744 (5) | -  5744 5 samples from LN     |
|               |               | - EGAD00001006144 (7) | -  Sorting vs spreading       |
+---------------+---------------+-----------------------+-------------------------------+
| S2            | 0             | N/A                   | N/A                           |
|               |               |                       |                               |
+---------------+---------------+-----------------------+-------------------------------+
| S3            | 14            | - EGAD00001004810 (14)| -  4810 has 11xJuxta-tumor    |
|               |               |                       | -  Ovarian                    |
+---------------+---------------+-----------------------+-------------------------------+
| S4            | 15            | - EGAD00001003808 (10)| -  3808 has 9xJuxta-tumor     |
|               |               | - EGAD00001005744 (5) | -  5744 5 samples from LN     |
+---------------+---------------+-----------------------+-----------------------+

With the juxta-tumour data, they got tumour and juxta-tumour data from the same patient. However, I have not been able to figure out whether they came from the same patient. Could probably use Optitype to determine HLA allele - match tumour and juxta tumour.

We also have scRNA-seq data for S1.

The data was processed using nf-core/rnaseq version `3.8.1` using the default parameters. STAR/Salmon were used for alignment/quantification.

We would expect our tumour-associated normal to be most like the S3 subtype (usually accumulate in juxta-tumours).

Combining RNA-sequencing datasets from different studies can be very challenging. We can expect batch effects to be present, so it might be possible to determine whether differences be observe are due to actual biological effects or technical artifacts. In addition, a recent study suggests that DESeq2 and edgeR (the most popular differential expression tools) experience large rates of false positives when used with large sample sizes [@Li2022]. One of the datasets (`EGAD00001006144`) was produced using stranded RNA-seq, whereas the other datasets were unstranded. This can lead to a lack of comparability of the datasets [@Zhao2020]. It may be necessary to drop this dataset from the analysis. All samples were prepared by poly(A) selection (use of oligo-dT).

## Preparation
### Create Sample File

Columns will be: Sample, Study, CAF_subtype, Tumor_Juxtatumor

*Here we will be combining data from 5 studies. To begin with, we will only include the metadata available for all studies (except for our unknown CAF subtype label). Breast cancer subtype is only available for certain studies and so is not included at this stage.*

There are also: ovarian cancer samples, EPCAM+ cells, samples prepared by spreading or spreading and samples from lymph nodes. For the time being, I will not consider them.

```{r message=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(stringr))
library(biomaRt)
suppressPackageStartupMessages(library(tximport))
library(DT)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
library(cowplot)
suppressPackageStartupMessages(library(PCAtools))
library(dplyr)
suppressPackageStartupMessages(library(SummarizedExperiment))
library(DESeq2)
suppressPackageStartupMessages(library(tximeta))
library(pheatmap)
library(RColorBrewer)
library(glmpca)
library(hexbin)
```

```{r read in metadata}
EGAD_4810 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001004810/delimited_maps/Run_Sample_meta_info.map", 
                        sep = ";")
EGAD_4810_cancers <- str_split_fixed(EGAD_4810$V1, pattern = "=", n = 2)[,2]
EGAD_4810_keep <- which(EGAD_4810_cancers == "BC")
EGAD_4810_filtered <- EGAD_4810[EGAD_4810_keep,]
EGAD_4810_meta <- data.frame(
  Sample = str_split_fixed(EGAD_4810_filtered$V4, pattern = "=", n = 2)[,2],
  Study = "EGAD00001004810",
  Subtype = "S3",
  Tumor_JuxtaTumor = tolower(str_split_fixed(EGAD_4810_filtered$V3, pattern = " ", n = 2)[,2]),
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001004810_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_3808 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001003808/meta_CAF-S1_S4_BC_47samples.txt", 
                        header = T, sep = "\t")
EGAD_3808_meta <- data.frame(
  Sample = EGAD_3808$Sample.Name,
  Study = "EGAD00001003808",
  Subtype = EGAD_3808$subset,
  Tumor_JuxtaTumor = EGAD_3808$Type, 
    directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001003808_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_6144 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001006144/meta_7samples.txt", 
                        header = T,sep = "\t")
EGAD_6144_meta <- data.frame(
  Sample = paste("CAF_Culture_", EGAD_6144$Sample.Name, sep = ""),
  Study = "EGAD00001006144",
  Subtype = "S1",
  Tumor_JuxtaTumor = "tumor",
    directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001006144_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_5744 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001005744/metaData_Pelon_et_al.txt", 
                        header =T, check.names = F)
EGAD_5744$Sample.Name <- gsub(pattern = "\\.", replacement = "-", x = EGAD_5744$Sample.Name )
EGAD_5744_filtered <- EGAD_5744[!(EGAD_5744$subset == "EPCAM+") & (EGAD_5744$Type == "T"),]
EGAD_5744_meta <- data.frame(
  Sample = EGAD_5744_filtered$Sample.Name,
  Study = "EGAD00001005744",
  Subtype = EGAD_5744_filtered$subset,
  Tumor_JuxtaTumor = "tumor",
    directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001005744_nfcore_results/star_salmon",
  row.names = 1
)
barkley_samples <- read.csv("/home/kevin/Documents/PhD/rna_seq_bc/metadata/reformat_samples.csv", 
                            header = T, row.names = "samples", check.names = F)
barkley_samples_meta <- data.frame(
  Sample = row.names(barkley_samples),
   Study = "InHouse",
  Subtype = "Unknown",
  Tumor_JuxtaTumor = ifelse(barkley_samples$Condition == "Tumour", "tumor", "juxtatumor"),
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/inhouse_data_nfcore_results_version_3_8_1/star_salmon",
  row.names = 1
)
metadata <- rbind.data.frame(EGAD_4810_meta, EGAD_3808_meta, 
                             EGAD_6144_meta, EGAD_5744_meta, barkley_samples_meta)
metadata$Tumor_JuxtaTumor <- gsub(x = metadata$Tumor_JuxtaTumor, pattern = "-", replacement = "")
metadata[1:5,]
```

### Read in data with tximeta and create DESeq object

Samples were processed with nf-core/rnaseq version `3.8.1`
Read in samples with tximport, deseqdataobject etc


```{r}
files <- file.path(metadata$directory, rownames(metadata), "quant.sf")
coldata <- data.frame(files, names=rownames(metadata), Study = metadata$Study, 
                      Subtype = metadata$Subtype, 
                      Tumor_JuxtaTumor = metadata$Tumor_JuxtaTumor,
                      stringsAsFactors=FALSE)
# tx2gene file for the gencode v31 file used in the analysis
tx2gene <- read_tsv("/home/kevin/Documents/PhD/references/tx2gene_gencode_v31.txt")
se <- tximeta(coldata, skipMeta=TRUE, txOut=FALSE, tx2gene=tx2gene)
```


```{r Apply common filters according to DESeq2 vignette to remove lowly expressed genes}
dds <- DESeqDataSet(se, design = ~ Study)
dim(dds)
# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds)) >= 10
# only keep rows (genes) for which keep is TRUE
dds <- dds[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds) >= 10) >= X
dds <- dds[keep,]
ntd <- normTransform(dds)
```

```{r Create DESeq object}
dds <- DESeq(dds)
```

### Data transformation

There are a number of options to choose from when normalising RNA-seq data, the main ones being:
 -  Take the log of the data and add a pseudocount.
 -  Variance stabilizing transformation (Anders and Huber 2010)
 -  Regularized logarithm transformation (Love, Huber, and Anders 2014)

The log+pseudocount approach tends to mean that lowly expressed genes have more of an effect. *Vst* and *rlog* bring these counts towards a central amount, making the data more homoskedastic. This allows them to be used in downstream processes which require homoskedastic data (e.g. PCA). The authors of DESeq2 recommend *vst* for large sample sizes such as ours as it is much faster than *rlog*


```{r}
# here, blind is FALSE as we don't want it to be blind to experimental design 
# recommended when transforming data for downstream analysis which will use the design information
vsd <- vst(dds, blind = FALSE)
```

```{r}
dds <- estimateSizeFactors(dds)

df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

There doesn't seem to much of a difference between the two methods of normalisation, only that the lowly expressed genes have been brought up to a minimum of ~4. 

### QC - PCA per study for outlier detection

```{r}
#studies <- levels(colData(vsd)$Study)
# write a function which takes in DESeq object & creates PCA of all studies
deseq_pca_studies <- function(dds_object){
  library(PCAtools)
  library(ggplot2)
  studies <- levels(colData(dds_object)$Study)
  pca_plots <- list()
  for (i in 1:length(studies)){
    study <- studies[i]
    samples <- colnames(dds_object)[which(colData(dds_object)$Study == study)]
    dds_object_study <- dds_object[,samples]
    stopifnot(dim(dds_object_study)[2] == length(samples))
    pca_plot_data <- plotPCA(dds_object_study, intgroup = c("Subtype", "Tumor_JuxtaTumor"), returnData = TRUE)
    percentVar <- round(100 * attr(pca_plot_data, "percentVar"))
    title <- paste("PCA plot of study:", study, sep = " ")
    #plot_labels <- rownames(pca_plot_data)
    #print(samples)
    #print(length(samples))
    pca_plot_data$samples <- rownames(pca_plot_data)
    pca_plot <- ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Subtype, shape = Tumor_JuxtaTumor, label = samples)) +
                geom_point(size =3) +
                geom_text(hjust="middle", vjust=1, data = subset(pca_plot_data, PC2 > 22)) +
                xlab(paste0("PC1: ", percentVar[1], "% variance")) +
                ylab(paste0("PC2: ", percentVar[2], "% variance")) +
                coord_fixed() +
                ggtitle(title)
    pca_plots[[i]] <- pca_plot
    #print(pca_plots[i])
  }
  #stopifnot(length(pca_plots) == length(studies))
  return(pca_plots)
}
```

```{r}
output <- deseq_pca_studies(vsd)
output
#output
```

```{r}
plotPCA(vsd, intgroup = c("Study", "Subtype"))

```


### Batch correction

The following graphs demonstrate the batch effects

```{r}
ann_colors = list(Study=c(EGAD00001004810 = "forestgreen",  EGAD00001003808 = "gold", EGAD00001006144 = "blue", EGAD00001005744 = "magenta3", InHouse = "black"),
                  Subtype = c(S1 = "dodgerblue4", S3 = "chartreuse1", S4 = "grey67", Unknown = "palevioletred"),
                  Tumor_JuxtaTumor = c(tumor = "royalblue", juxtatumor = "red")
                  )
```

```{r Generate distance matrix + generate heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
sampleDists <- dist(t(assay(vsd)))
# don't want files column to be on heatmap
subset_coldata <- subset(coldata, select = -c(files))
sampleDistMatrix <- as.matrix( sampleDists )
rownames(subset_coldata) <- subset_coldata$names
subset_coldata$names <- NULL

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```



```{r}
plotPCA(vsd, intgroup = c("Study", "Subtype"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup = c("Study", "Subtype"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData, aes(x = PC1, y = PC2, color = Study, shape = Subtype)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")
```

Next step is to look at batch correction.

#### Method 1: Limma's remove batch effect function

```{r message=FALSE}
library(limma)
```

```{r create copies of vsd to avoid confusion}
vsd_remove_batch_intercept <- vsd
vsd_remove_batch_tumor_juxta_subtype <- vsd
vsd_remove_batch_tumor_juxta_only <- vsd
```


```{r removeBatchEffect model matrix with only an intercept term}
mat <- assay(vsd_remove_batch_intercept)
mm <- model.matrix(~1, colData(vsd_remove_batch_intercept))
mat <- limma::removeBatchEffect(mat, batch = vsd_remove_batch_intercept$Study, design = mm)
assay(vsd_remove_batch_intercept) <- mat
pca_batch_correct_intercept <- plotPCA(vsd, intgroup = c("Subtype","Study")) + 
  aes(x = PC1, y = PC2, color = colData(vsd_remove_batch_intercept)$Subtype, shape =colData(vsd_remove_batch_intercept)$Study ) +
  labs(color='Subpopulation', shape = "Study") +
  ggtitle("PCA of VST data batch corrected using \nlimma::removeBatchEffect with intercept term only")
pca_batch_correct_intercept
```

```{r removeBatchEffect model matrix with only an intercept term, heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
# same template as previous heatmap
sampleDists <- dist(t(assay(vsd_remove_batch_intercept)))
sampleDistMatrix <- as.matrix( sampleDists )

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```

Limma's removeBatchEffect function requires a design matrix as input, this is the "treatment conditions" we wish to preserve. It is usually the design matrix with all experimental factors other than batch effects. The ideal scenario would be to include Subtype in this matrix. However, this treats `Unknown` as its own subtype, and so will preserve differences between the InHouse samples and the other samples, as seen in the below PCA plot. This is contrary to what we want when assigning our samples to a cluster.

```{r removeBatchEffect tumor juxta + subtype model matrix}
mat <- assay(vsd_remove_batch_tumor_juxta_subtype)
# create model matrix
mm <- model.matrix(~Tumor_JuxtaTumor+Subtype, colData(vsd_remove_batch_tumor_juxta_subtype))
mat <- limma::removeBatchEffect(mat, batch = vsd_remove_batch_tumor_juxta_subtype$Study, design = mm)
assay(vsd_remove_batch_tumor_juxta_subtype) <- mat

pca_batch_correct_tumor_juxta_subtype <- plotPCA(vsd_remove_batch_tumor_juxta_subtype, intgroup = c("Subtype","Study")) + 
  aes(x = PC1, y = PC2, color = colData(vsd_remove_batch_tumor_juxta_subtype)$Subtype, shape =colData(vsd_remove_batch_tumor_juxta_subtype)$Study ) +
  labs(color='Subpopulation', shape = "Study") +
  ggtitle("PCA of VST data batch corrected using \nlimma::removeBatchEffect with Tumor_JuxtaTumor\nand Subtype in model")
pca_batch_correct_tumor_juxta_subtype
```

```{r removeBatchEffect model matrix with Study+Subpopulation, heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
# same template as previous heatmap

sampleDists <- dist(t(assay(vsd_remove_batch_tumor_juxta_subtype)))
sampleDistMatrix <- as.matrix( sampleDists )

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```

We can see from the PCA and heatmap above that including condition (tumour-juxtatumour) and subtype in our model matrix leads to good separation of the subtypes, and a lack of clustering by batch. The problem with this approach is that it treats our `Unknown` subtype as its own subtype, which means it will cluster on its own no matter what we do. Is there a way around this? Frozen surrogate variable analysis could be an option, but this is usually used on microarray data, I don't know if it can be used on RNA-sequencing data.

```{r removeBatchEffect tumor juxta only matrix}
mat <- assay(vsd)
mm <- model.matrix(~Tumor_JuxtaTumor, colData(vsd))
mat_batch_removed_limma <- limma::removeBatchEffect(mat, batch = vsd$Study, design = mm)
```

In the PCA plot above, we have not told the `removeBatchEffect` function about our known subtypes, only whether the samples were taken from Tumor or Juxta-Tumor. It does not know to preserve differences between subpopulations when removing batch effects. Less variance being explained by PC1 than in the first scenario (39% vs 68%). In the first PCA plot, our in-house samples of unknown subtype cluster together on their own. 


```{r run BatchQC on our data}
#mat <- assay(vsd)
#modmatrix <- model.matrix(~as.factor(Tumor_JuxtaTumor), colData(vsd))
#batchQC(mat, batch=coldata$Study, condition= coldata$Subtype, 
      #  report_file = "batchqc_caf_data_not_corrected.html",
       # report_dir = ".", view_report = FALSE, interactive = FALSE
      #  )
```


```{r}
coldata_pca <- coldata
rownames(coldata_pca) <- coldata_pca$names
coldata_pca$names <- NULL
coldata_pca$files <- NULL
p <- pca(mat, metadata = coldata_pca)

```

```{r}
ggplot(p$rotated, aes(x = PC1, y = PC2, color = p$metadata$Study, shape = p$metadata$Subtype)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  labs(color='Study', shape = "Subpopulation") +
  ggtitle("PCA with transormed data after batch correction")
```

```{r Batch correction using Combat-Seq}
batch <- coldata$Study
mm <- model.matrix(~Tumor_JuxtaTumor, colData(vsd))
```



## Assigning in-house samples to a CAF subtype using K-nearest neighbours

1. Split mat into our known (training) and unknown (testing) subpopulations
2. Run KNN


```{r create separate variables for known and unknown CAF subtypes}
mat_t <- t(mat)
mat_known <- mat_t[coldata$names[which(coldata$Subtype != "Unknown")],]
coldata_known <- coldata[coldata$Subtype != "Unknown",]
mat_unknown <- mat_t[coldata$names[which(coldata$Subtype == "Unknown")],]
coldata_unknown <- coldata[coldata$Subtype == "Unknown",]
```

```{r}
library(class)
```

```{r}
##create a random number equal 90% of total number of rows
 ran <- sample(1:nrow(mat_known),0.9 * nrow(mat_known))
 ##training dataset extracted
 mat_train <- mat_known[ran,]
 
 ##test dataset extracted
 mat_test <- mat_known[-ran,]
```

```{r}
# these are our labels
caf_target_category <- coldata_known[ran,4]
caf_test_category <- coldata_known[-ran,4]
```

```{r run KNN}
pr <- knn(mat_train,mat_test,cl=caf_target_category,k=21)
```

```{r}
 tab <- table(pr,caf_test_category)
tab
```
```{r}
accuracy <- function(x){sum(diag(x)/(sum(rowSums(x)))) * 100}
 accuracy(tab)
```
```{r}
  outputs <- c()
for (i in 1:50){
  pr <- knn(mat_train,mat_test,cl=caf_target_category,k=i)
  number <- i
  tab <- table(pr,caf_test_category)
  accuracy_out <- accuracy(tab)
  outputs <- c(outputs, number = accuracy_out)
}
```

```{r}
plot(outputs)
```

```{r}
  prediction <- knn(mat_train,mat_unknown,cl=caf_target_category,k=13)
prediction
```
They are all predicted to be S1 using this initial application of the algorithm.
Possibly use interquartile range to improve performance?

## References
