---
title: "CAF subtype analysis"
author: "Kevin Ryan"
date: "`r Sys.time()`"
bibliography: citations.bib
output: 
  github_document:
     toc: true
     toc_depth: 3 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Cancer-associated fibroblasts (CAFs) are a heterogeneous cell type found in the tumour microenvironment. They have a wide array of functions, and tend to be immunosuppressive and cancer-promoting. There have been many attempts to characterise subpopulations of CAFs, with much transcriptomic analysis being carried out in the Mechta-Grigoriou lab in Institut Curie. They have identified 4 'subpopulations' which can be separated based on the expression of different markers:

 -  S1: FAP^High^, CD29^Med-High^, α^SMAHigh^, PDPN^High^, PDGFRβ^High^
 -  S2: FAP^Neg^, CD29^Low^, αSMANeg-^Low^, PDPN^Low^, PDGFRβ^Low^
 -  S3: FAP^Neg-Low^, CD29^Med^, αSMA^Neg-Low^, PDPN^Low^, PDGFRβ^Low-Med^
 -  S4: FAP^Low-Med^, CD29^High^, αSMA^High^, PDPN^Low^, PDGFRβ^Med^

[@Pelon2020]

FACS gating strategies can be used to isolate these various subpopulations. The Mechta-Grigoriou group have done this and have generated bulk RNA-sequencing data for the S1, S3 and S4 subpopulations. They generated scRNA-sequencing data for the S1 subpopulation. This data was deposited on the European Genome Phenome Archive, and was accessed via a Data Transfer Agreement.

The following summarises the data obtained:

+---------------+---------------+-----------------------+-------------------------------+
| Subpopulation | Total samples | Studies (Samples)     |  Notes                        |
+===============+===============+=======================+===============================+
| S1            |  28           | - EGAD00001003808 (16)| -  3808 has 12xJuxta-tumor    |
|               |               | - EGAD00001005744 (5) | -  5744 5 samples from LN     |
|               |               | - EGAD00001006144 (7) | -  Sorting vs spreading       |
+---------------+---------------+-----------------------+-------------------------------+
| S2            | 0             | N/A                   | N/A                           |
|               |               |                       |                               |
+---------------+---------------+-----------------------+-------------------------------+
| S3            | 14            | - EGAD00001004810 (14)| -  4810 has 11xJuxta-tumor    |
|               |               |                       | -  Ovarian                    |
+---------------+---------------+-----------------------+-------------------------------+
| S4            | 15            | - EGAD00001003808 (10)| -  3808 has 9xJuxta-tumor     |
|               |               | - EGAD00001005744 (5) | -  5744 5 samples from LN     |
+---------------+---------------+-----------------------+-----------------------+

With the juxta-tumour data, they got tumour and juxta-tumour data from the same patient. However, I have not been able to figure out whether they came from the same patient. Could probably use Optitype to determine HLA allele - match tumour and juxta tumour.

We also have scRNA-seq data for S1.

It is likely that sorting the cells using FACS alters the transcriptional properties of the cells compared to if they are separated using spreading approches, as is seen in study `EGAD00001006144`. This is something that we will have to keep in mind.

The data was processed using nf-core/rnaseq version `3.8.1` using the default parameters. STAR/Salmon were used for alignment/quantification.

We would expect our tumour-associated normal to be most like the S3 subtype (usually accumulate in juxta-tumours).

Combining RNA-sequencing datasets from different studies can be very challenging. We can expect batch effects to be present, so it might be possible to determine whether differences be observe are due to actual biological effects or technical artifacts. In addition, a recent study suggests that DESeq2 and edgeR (the most popular differential expression tools) experience large rates of false positives when used with large sample sizes [@Li2022]. One of the datasets (`EGAD00001006144`) was produced using stranded RNA-seq, whereas the other datasets were unstranded. This can lead to a lack of comparability of the datasets [@Zhao2020]. It may be necessary to drop this dataset from the analysis. All samples were prepared by poly(A) selection (use of oligo-dT).

## Preparation
### Create Sample File

Columns will be: Sample, Study, CAF_subtype, Tumor_Juxtatumor

*Here we will be combining data from 5 studies. To begin with, we will only include the metadata available for all studies (except for our unknown CAF subtype label). Breast cancer subtype is only available for certain studies and so is not included at this stage.*

There are also: ovarian cancer samples, EPCAM+ cells, samples prepared by spreading or spreading and samples from lymph nodes. For the time being, I will not consider them.

```{r message=FALSE, echo=FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(stringr))
library(biomaRt)
suppressPackageStartupMessages(library(tximport))
library(DT)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
library(cowplot)
suppressPackageStartupMessages(library(PCAtools))
library(dplyr)
suppressPackageStartupMessages(library(SummarizedExperiment))
library(DESeq2)
suppressPackageStartupMessages(library(tximeta))
library(pheatmap)
library(RColorBrewer)
library(glmpca)
library(hexbin)
suppressPackageStartupMessages(library(IHW))
httr::set_config(httr::config(ssl_verifypeer = FALSE))
```

```{r read in metadata}
EGAD_4810 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001004810/delimited_maps/Run_Sample_meta_info.map", 
                        sep = ";")
EGAD_4810_cancers <- str_split_fixed(EGAD_4810$V1, pattern = "=", n = 2)[,2]
EGAD_4810_keep <- which(EGAD_4810_cancers == "BC")
EGAD_4810_filtered <- EGAD_4810[EGAD_4810_keep,]
EGAD_4810_meta <- data.frame(
  Sample = str_split_fixed(EGAD_4810_filtered$V4, pattern = "=", n = 2)[,2],
  Study = "EGAD00001004810",
  Subtype = "S3",
  Tumor_JuxtaTumor = tolower(str_split_fixed(EGAD_4810_filtered$V3, pattern = " ", n = 2)[,2]),
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001004810_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_3808 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001003808/meta_CAF-S1_S4_BC_47samples.txt", 
                        header = T, sep = "\t")
EGAD_3808_meta <- data.frame(
  Sample = EGAD_3808$Sample.Name,
  Study = "EGAD00001003808",
  Subtype = EGAD_3808$subset,
  Tumor_JuxtaTumor = EGAD_3808$Type, 
    directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001003808_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_6144 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001006144/meta_7samples.txt", 
                        header = T,sep = "\t")
EGAD_6144_meta <- data.frame(
  Sample = paste("CAF_Culture_", EGAD_6144$Sample.Name, sep = ""),
  Study = "EGAD00001006144",
  Subtype = "S1",
  Tumor_JuxtaTumor = "tumor",
    directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001006144_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_5744 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001005744/metaData_Pelon_et_al.txt", 
                        header =T, check.names = F)
EGAD_5744$Sample.Name <- gsub(pattern = "\\.", replacement = "-", x = EGAD_5744$Sample.Name )
EGAD_5744_filtered <- EGAD_5744[!(EGAD_5744$subset == "EPCAM+") & (EGAD_5744$Type == "T"),]
EGAD_5744_meta <- data.frame(
  Sample = EGAD_5744_filtered$Sample.Name,
  Study = "EGAD00001005744",
  Subtype = EGAD_5744_filtered$subset,
  Tumor_JuxtaTumor = "tumor",
    directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001005744_nfcore_results/star_salmon",
  row.names = 1
)
barkley_samples <- read.csv("/home/kevin/Documents/PhD/rna_seq_bc/metadata/reformat_samples.csv", 
                            header = T, row.names = "samples", check.names = F)
barkley_samples_meta <- data.frame(
  Sample = row.names(barkley_samples),
   Study = "InHouse",
  Subtype = "Unknown",
  Tumor_JuxtaTumor = ifelse(barkley_samples$Condition == "Tumour", "tumor", "juxtatumor"),
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/inhouse_data_nfcore_results_version_3_8_1/star_salmon",
  row.names = 1
)
metadata <- rbind.data.frame(EGAD_4810_meta, EGAD_3808_meta, 
                             EGAD_6144_meta, EGAD_5744_meta, barkley_samples_meta)
metadata$Tumor_JuxtaTumor <- gsub(x = metadata$Tumor_JuxtaTumor, pattern = "-", replacement = "")
metadata[1:5,]
write.table(metadata, file = "/home/kevin/Documents/PhD/subtypes/caf-subtype-analysis/metadata_caf_subtypes.txt", quote = F, sep = "\t", row.names = T)
metadata_no_inhouse <- rbind.data.frame(EGAD_4810_meta, EGAD_3808_meta, 
                             EGAD_6144_meta, EGAD_5744_meta)
metadata_no_inhouse$Tumor_JuxtaTumor <- gsub(x = metadata_no_inhouse$Tumor_JuxtaTumor, pattern = "-", replacement = "")

```

### Read in data with tximeta and create DESeq object

Samples were processed with nf-core/rnaseq version `3.8.1`
Read in samples with tximport, deseqdataobject etc


```{r}
files <- file.path(metadata$directory, rownames(metadata), "quant.sf")
coldata <- data.frame(files, names=rownames(metadata), Study = metadata$Study, 
                      Subtype = metadata$Subtype, 
                      Tumor_JuxtaTumor = metadata$Tumor_JuxtaTumor,
                      stringsAsFactors=FALSE)
# tx2gene file for the gencode v31 file used in the analysis
tx2gene <- read_tsv("/home/kevin/Documents/PhD/references/tx2gene_gencode_v31.txt")
se <- tximeta(coldata, skipMeta=TRUE, txOut=FALSE, tx2gene=tx2gene)
```

```{r read in data without inhouse}
files_no_inhouse <- file.path(metadata_no_inhouse$directory, rownames(metadata_no_inhouse), "quant.sf")
coldata_no_inhouse <- data.frame(files = files_no_inhouse, names=rownames(metadata_no_inhouse), Study = metadata_no_inhouse$Study, 
                      Subtype = metadata_no_inhouse$Subtype, 
                      Tumor_JuxtaTumor = metadata_no_inhouse$Tumor_JuxtaTumor,
                      stringsAsFactors=FALSE)
# tx2gene file for the gencode v31 file used in the analysis
#tx2gene <- read_tsv("/home/kevin/Documents/PhD/references/tx2gene_gencode_v31.txt")
se_no_inhouse <- tximeta(coldata = coldata_no_inhouse, skipMeta=TRUE, txOut=FALSE, tx2gene=tx2gene)
```

```{r Apply common filters according to DESeq2 vignette to remove lowly expressed genes}
# do no design for the time being, Subtype + Batch gives error - model matrix not full rank
# this function stores input values, intermediate calculations and results of DE analysis - makes counts non-negative integers
dds_no_inhouse <- DESeqDataSet(se_no_inhouse, design = ~1)
# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds_no_inhouse)) >= 10
# only keep rows (genes) for which keep is TRUE
dds_no_inhouse <- dds_no_inhouse[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds_no_inhouse) >= 10) >= X
dds_no_inhouse <- dds_no_inhouse[keep,]
dds_no_inhouse_copy <- dds_no_inhouse
ntd <- normTransform(dds_no_inhouse)
```

```{r}
# do no design for the time being, Subtype + Batch gives error - model matrix not full rank
# this function stores input values, intermediate calculations and results of DE analysis - makes counts non-negative integers
dds <- DESeqDataSet(se, design = ~1)
# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds)) >= 10
# only keep rows (genes) for which keep is TRUE
dds <- dds[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds) >= 10) >= X
dds <- dds[keep,]
ntd <- normTransform(dds)
```



```{r}
suppressPackageStartupMessages(library(sva))
dds_no_inhouse <- DESeq(dds_no_inhouse)
dat  <- counts(dds_no_inhouse, normalized = TRUE)
idx  <- rowMeans(dat) > 1
dat  <- dat[idx, ]
# we have subtype and tumour-juxtatomour in the model matrix
mod  <- model.matrix(~ Subtype + Tumor_JuxtaTumor, colData(dds_no_inhouse))
mod0 <- model.matrix(~   1, colData(dds_no_inhouse))
svseq <- svaseq(dat, mod, mod0, n.sv = 4)
```
```{r}
#svseq$sv
par(mfrow = c(4, 1), mar = c(3,5,3,1))
for (i in 1:4) {
  stripchart(svseq$sv[, i] ~ dds_no_inhouse$Study, vertical = TRUE, main = paste0("SV", i))
  abline(h = 0)
 }
```

```{r}
dds_no_inhouse$SV1 <- as.factor(svseq$sv[,1])
dds_no_inhouse$SV2 <- as.factor(svseq$sv[,2])
dds_no_inhouse$SV3 <- as.factor(svseq$sv[,3])
dds_no_inhouse$SV4 <- as.factor(svseq$sv[,4])
dds_no_inhouse$Subtype <- as.factor(dds_no_inhouse$Subtype)
dds_no_inhouse$Tumor_JuxtaTumor <- as.factor(dds_no_inhouse$Tumor_JuxtaTumor)
design(dds_no_inhouse) <- ~ 1

```

```{r}
dds_no_inhouse_deseq <- DESeq(dds_no_inhouse)
```




```{r}
metadata_subtype_study <- data.frame(metadata$Study, metadata$Subtype)
```


```{r Create DESeq object}
dds <- DESeq(dds)
```

```{r get HGNC symbols for genes}
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host="useast.ensembl.org")
ensembl_ids <- rownames(assay(dds))
ensembl_ids <- str_split_fixed(ensembl_ids, pattern = "\\.", n = 2)[,1]
info <- getBM(attributes=c("hgnc_symbol", "ensembl_gene_id"),
                  filters = "ensembl_gene_id",
                  values = ensembl_ids,
                  mart = mart,
                  useCache=FALSE)

#rownames(info) <- info$hgnc_symbol
```
Source: GeneCards
 -  ITGB1 (CD29) is an integrin, integrins are involved in cell-cell and cell-ECM adhesion. It is expressed in 3 of the 4 subpopulations (not S2).
 -  FAP is Fibroblast activation protein, a serine proteinase, selectively expressed in reactive stromal fibroblasts in epithelial cancers, expression should be highest in S1 population. Roles include: tissue remodelling, fibrosis, wound healing.
 -  α^SMA - ACTA2, alpha smooth muscle actin is involved in cell motility, structure, integrity and signalling. Its expression is highest in S1 and S4 CAFs.
 -  PDPN (podoplanin) increases cell motility. It should only be expressed in S1 cells.
 -  PDGFRβ (Platelet Derived Growth Factor Receptor Beta) is a tyrosine kinase receptor for platelet-derived growth factors. They stimulate mitosis of cells of mesenchymal origin. Expression should be highest in S1, some expression in S3 and S4.
 
There are other possible genes which could be used to distinguish them through manual curation: FSP1, CAV1, DPP4
 
```{r Plotting functions}
# Function to take in dds object, ensembl - hgnc info and genes interest and output df for plotting
create_annotated_df_for_plotting <- function(dds, info, genes_interest){
  library(stringr)
  # put check for type of gene symbol in dds object, assuming ensembl gene version atm
  # assume info have column called hgnc_symbol
  info <- info[!duplicated(info$hgnc_symbol),]
  rownames(info) <- info$hgnc_symbol
  # assuming genes interest are in info, put in check for this to deal with missing gene symbol
  genes_interest_info <- info[genes_interest,] 
  # again, here assuming ENSG1234.1 etc
  ensembl_genes_dds <- str_split_fixed(rownames(assay(dds)), pattern = "\\.", n = 2)[,1]
  ids <- which(ensembl_genes_dds %in% genes_interest_info$ensembl_gene_id)
  dds_genes_interest <- dds[ids,]
  dds_genes_interest_exprs <- data.frame(assay(dds_genes_interest))
  metadata <- as.data.frame(colData(dds))
  colnames(dds_genes_interest_exprs) <- rownames(metadata)
  rownames(dds_genes_interest_exprs) <- genes_interest
  dds_genes_interest_exprs_t <- as.data.frame(t(dds_genes_interest_exprs))
  dds_genes_interest_exprs_t$Subpopulation <- metadata$Subtype
  dds_genes_interest_exprs_t$Study <- metadata$Study
  dds_genes_interest_exprs_t$Tumor_JuxtaTumor <- metadata$Tumor_JuxtaTumor
  return(dds_genes_interest_exprs_t)
}

caf_plot_tumour_juxtatumour <- function(annotated_df_for_plotting, gene_of_interest, label_yaxis = gene_of_interest){
  if (is.null(label_yaxis)){
    label_yaxis <- gene_of_interest
  }
  annotated_df_for_plotting$gene_of_interest <- annotated_df_for_plotting[, gene_of_interest]
  ggplot(annotated_df_for_plotting, aes(Subpopulation, gene_of_interest, colour = Tumor_JuxtaTumor)) +
  geom_point(size = 1,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  scale_y_continuous(gene_of_interest) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank())  +
    ylab(label_yaxis)  #, axis.title.x = element_blank())
}

caf_plot_study <- function(annotated_df_for_plotting, gene_of_interest, label_yaxis = gene_of_interest){
  if (is.null(label_yaxis)){
    label_yaxis <- gene_of_interest
  }
  annotated_df_for_plotting$gene_of_interest <- annotated_df_for_plotting[, gene_of_interest]
  ggplot(annotated_df_for_plotting, aes(Subpopulation, gene_of_interest, colour = Study)) +
  geom_point(size = 1,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  scale_y_continuous(gene_of_interest) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank()) +
    ylab(label_yaxis)#
}
```

```{r Prepare vsd object for plotting genes of interest}
library(ggpubr)
vsd <- vst(dds, blind = FALSE)
genes_interest <- c("FAP", "ITGB1", "ACTA2", "PDPN", "PDGFRB")
dds_genes_interest_exprs_t <- create_annotated_df_for_plotting(vsd, info = info, genes_interest = genes_interest)
df_for_plotting <- create_annotated_df_for_plotting(dds = dds, info = info, genes_interest = genes_interest)
caf_plot_study(df_for_plotting, gene_of_interest = "FAP")
caf_plot_tumour_juxtatumour(df_for_plotting, gene_of_interest = "FAP")
plots_out_study <- lapply(FUN = caf_plot_study, X = genes_interest, annotated_df_for_plotting = df_for_plotting)
plots_out_subpop <- lapply(FUN = caf_plot_tumour_juxtatumour, X = genes_interest, annotated_df_for_plotting = df_for_plotting)
ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_subpop <- ggarrange(plotlist = plots_out_subpop, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"))
ggar_obj_subpop_annotated <- annotate_figure(ggar_obj_subpop, bottom = text_grob("CAF Subpopulation"))
ggar_obj_study_annotated
ggar_obj_subpop_annotated
```

```{r Plot some other genes of interest}
genes_interest <- c("CXCL12","TNFSF4","PDCD1LG2", "CD276", "NT5E", "DPP4", "CAV1", "ATL1")
df_for_plotting <- create_annotated_df_for_plotting(dds = dds, info = info, genes_interest = genes_interest)
plots_out_study <- lapply(FUN = caf_plot_study, X = genes_interest, annotated_df_for_plotting = df_for_plotting)
plots_out_subpop <- lapply(FUN = caf_plot_tumour_juxtatumour, X = genes_interest, annotated_df_for_plotting = df_for_plotting)
ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_subpop <- ggarrange(plotlist = plots_out_subpop, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"))
ggar_obj_subpop_annotated <- annotate_figure(ggar_obj_subpop, bottom = text_grob("CAF Subpopulation"))
ggar_obj_study_annotated
ggar_obj_subpop_annotated
```

```{r Plot haemopoietic, epithelial and endothelial marker - should be low}
# plotting original counts here, transformation 
genes_interest <- c("PTPRC", "EPCAM", "PECAM1")
genes_interest_common_name <- c("CD45", "EPCAM", "CD31")
df_for_plotting <- create_annotated_df_for_plotting(dds = dds, info = info, genes_interest = genes_interest)
plots_out_study <- lapply(FUN = caf_plot_study, X = genes_interest, annotated_df_for_plotting = df_for_plotting, label_yaxis = genes_interest_common_name)
plots_out_subpop <- lapply(FUN = caf_plot_tumour_juxtatumour, X = genes_interest, annotated_df_for_plotting = df_for_plotting, label_yaxis = genes_interest_common_name)
ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_subpop <- ggarrange(plotlist = plots_out_subpop, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"))
ggar_obj_subpop_annotated <- annotate_figure(ggar_obj_subpop, bottom = text_grob("CAF Subpopulation"))
ggar_obj_study_annotated
ggar_obj_subpop_annotated
```

The low to zero counts of these genes are what we would expect.

```{r}
#fap_plot <- ggplot(dds_genes_interest_exprs_t, aes(x = Subpopulation, y = FAP, color = Tumor_JuxtaTumor, shape = Study)) +
 # geom_point(size = 0.75,  # reduce point size to minimize overplotting 
  #  position = position_jitter(
   #   width = 0.15,  # amount of jitter in horizontal direction
    #  height = 0     # amount of jitter in vertical direction (0 = none)
  #  )
#  ) +
#  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#panel.background = element_blank(), axis.line = element_line(colour = "black"))

itgb1_plot <- ggplot(dds_genes_interest_exprs_t, aes(x = Subpopulation, y = ITGB1, color = Tumor_JuxtaTumor, shape = Study)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

acta2_plot <- ggplot(dds_genes_interest_exprs_t, aes(x = Subpopulation, y = ACTA2, color = Tumor_JuxtaTumor, shape = Study)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

pdpn_plot <- ggplot(dds_genes_interest_exprs_t, aes(x = Subpopulation, y = PDPN, color = Tumor_JuxtaTumor, shape = Study)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

pdgfrb_plot <- ggplot(dds_genes_interest_exprs_t, aes(x = Subpopulation, y = PDGFRB, color = Tumor_JuxtaTumor, shape = Study)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),
legend.key.size = unit(0.25, 'cm'),
legend.title = element_text(size=10), #change legend title font size

)

#p <- plot_grid(fap_plot + theme(legend.position = "none"), 
 #              itgb1_plot + theme(legend.position = "none") , 
  #             acta2_plot + theme(legend.position = "none"), 
   #            pdpn_plot + theme(legend.position = "none"), 
    #           pdgfrb_plot, #+ theme(legend.position = "none"), 
     #          ncol=3, 
      #         labels=LETTERS[1:5])
#legend <- get_legend(
  # create some space to the left of the legend
 # fap_plot + theme(legend.box.margin = margin(0, 0, 0,12))
#)
# look again at what this is
#title <- ggdraw() + draw_label("Unnormalised expression of marker genes\naccording to CAF subpopulation", fontface='bold')
#plot_grid(title, p, legend, ncol=1, rel_heights=c(0.1, 0.5), rel_widths = 1, 1) # rel_heights values control title margins
#plot_grid(title, p, ncol=1, rel_heights=c(0.1, 0.5)) # rel_heights values control title margins

```

```{r}
library(ggpubr)
# input table - row sample, col genes

#caf_plot_function(dds_genes_interest_exprs_t, gene_of_interest = "FAP")
#plots_out <- lapply(FUN = caf_plot_function_tumour_juxtatumour, X = genes_interest, input_table = dds_genes_interest_exprs_t)
#plots_out
#ggar_obj <- ggarrange(plotlist = plots_out, common.legend = TRUE) # rel_heights values control title margins
#ggar_obj_annotated <- annotate_figure(ggar_obj, bottom = text_grob("CAF Subpopulation"))
#ggar_obj_annotated
```

```{r Define plotting functions}
caf_plot_function_study <- function(input_table, gene_of_interest){
  library(ggplot2)
  input_table$gene_of_interest <- input_table[, gene_of_interest]
  ggplot(input_table, aes(Subpopulation, gene_of_interest, colour = Study)) +
  geom_point(size = 1,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  scale_y_continuous(gene_of_interest) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank())
}

caf_plot_function_tumour_juxtatumour <- function(input_table, gene_of_interest){
  library(ggplot2)
  input_table$gene_of_interest <- input_table[, gene_of_interest]
  ggplot(input_table, aes(Subpopulation, gene_of_interest, colour = Tumor_JuxtaTumor)) +
  geom_point(size = 1,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  scale_y_continuous(gene_of_interest) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"), axis.title.x = element_blank())
}

```

```{r}
library(ggpubr)
genes_interest <- c("FAP", "ITGB1", "ACTA2", "PDPN", "PDGFRB")
plots_out <- lapply(FUN = caf_plot_function_study, X = genes_interest, input_table = dds_genes_interest_exprs_t)
ggar_obj <- ggarrange(plotlist = plots_out, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_annotated <- annotate_figure(ggar_obj, bottom = text_grob("CAF Subpopulation"))
ggar_obj_annotated
```

```{r Make plot for vsd data}
#vsd <- vst(dds, blind = FALSE)
info_unique <- info[!duplicated(info$hgnc_symbol),]
rownames(info_unique) <- info_unique$hgnc_symbol
genes_interest_info <- info_unique[genes_interest,] 
ensembl_genes_vsd <- str_split_fixed(rownames(assay(vsd)), pattern = "\\.", n = 2)[,1]
#dds_replicated <- dds
ids <- which(ensembl_genes_vsd %in% genes_interest_info$ensembl_gene_id)
vsd_genes_interest <- vsd[ids,]
vsd_genes_interest_exprs <- data.frame(assay(vsd_genes_interest))
colnames(vsd_genes_interest_exprs) <- rownames(metadata)
rownames(vsd_genes_interest_exprs) <- genes_interest
vsd_genes_interest_exprs_t <- as.data.frame(t(vsd_genes_interest_exprs))
vsd_genes_interest_exprs_t$Subpopulation <- metadata$Subtype
vsd_genes_interest_exprs_t$Study <- metadata$Study
vsd_genes_interest_exprs_t$Tumor_JuxtaTumor <- metadata$Tumor_JuxtaTumor
fap_plot <- ggplot(vsd_genes_interest_exprs_t, aes(x = Subpopulation, y = FAP)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))
itgb1_plot <- ggplot(vsd_genes_interest_exprs_t, aes(x = Subpopulation, y = ITGB1)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

acta2_plot <- ggplot(vsd_genes_interest_exprs_t, aes(x = Subpopulation, y = ACTA2)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

pdpn_plot <- ggplot(vsd_genes_interest_exprs_t, aes(x = Subpopulation, y = PDPN)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

pdgfrb_plot <- ggplot(vsd_genes_interest_exprs_t, aes(x = Subpopulation, y = PDGFRB)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"))

p <- plot_grid(fap_plot, itgb1_plot, acta2_plot, pdpn_plot, pdgfrb_plot, ncol=3, labels=LETTERS[1:5])
title <- ggdraw() + draw_label("VSD normalised expression of marker genes\naccording to CAF subpopulation", fontface='bold')
plot_grid(title, p, ncol=1, rel_heights=c(0.1, 0.5)) # rel_heights values control title margins
```

```{r}
plots_out <- lapply(FUN = caf_plot_function_study, X = genes_interest, input_table = vsd_genes_interest_exprs_t)
ggar_obj <- ggarrange(plotlist = plots_out, common.legend = TRUE, legend = "right") # rel_heights values control title margins
title <- ggdraw() + draw_label("VSD normalised expression of marker genes\naccording to CAF subpopulation", fontface='bold')
ggar_obj_annotated <- annotate_figure(ggar_obj, bottom = text_grob("CAF Subpopulation"), top = text_grob("VSD normalised expression of marker genes\naccording to CAF subpopulation", color = "black", face = "bold", size = 10))
ggar_obj_annotated

plots_out <- lapply(FUN = caf_plot_function_tumour_juxtatumour, X = genes_interest, input_table = vsd_genes_interest_exprs_t)
ggar_obj <- ggarrange(plotlist = plots_out, common.legend = TRUE, legend = "right") # rel_heights values control title margins
title <- ggdraw() + draw_label("VSD normalised expression of marker genes\naccording to CAF subpopulation", fontface='bold')
ggar_obj_annotated <- annotate_figure(ggar_obj, bottom = text_grob("CAF Subpopulation"), top = text_grob("VSD normalised expression of marker genes\naccording to CAF subpopulation", color = "black", face = "bold", size = 10))
ggar_obj_annotated
```




```{r Expression of genes used to exclude haemopoietic, epithelial and endothelial cells}

```

```{r}
library(purrr)
#summarised_data <- vsd_genes_interest_exprs_t %>% split(.$Subtype) %>% map(summary)
#summarised_data$S1
```
Per Gene, do the data line up with what we would expect?
 -  FAP, S1 = high, would expect it to be lower in S3 than S4, they are roughly the same. Unknown subtype, high, tight confidence
 -  ITGB1 (CD29), very similar across all subtypes. This is as expected - it is thought to not be expressed only in S2 which we don't have.
 -  ACTA2 (alphaSMA), very similar across all subpopulations. We would expect the expression to be lowest in the S3 subpopulation but we don't see that here.
 -  PDPN - expression highest in unknown samples, similar but lower in other subpopulations. Slightly higher in S1 than S3 and S4 - supposed to be expressed only in S1! Post-transcriptional regulation??? Seems to undergo post-translational regulation - https://www.spandidos-publications.com/10.3892/ijo.2013.1887#b53-ijo-42-06-1849 it is possible that differences won't be seen here on the mRNA level.
 -  PDGFRB - highest in S1, as expected. Expression in our unknown similar to S3 and S4.


 -  Carry out batch correction on reduced dataset, tumour_juxtatumour and subtype as model matrix
 -  Plot genes
 -  DE analysis
 
```{r}
library(sva)
counts_matrix <- assay(dds_no_inhouse_copy)
batch <- metadata_no_inhouse$Study
covariates <- metadata_no_inhouse$Tumor_JuxtaTumor
  #Subpopulation = metadata_no_inhouse$Subtype)
                        
adjusted <- ComBat_seq(counts = counts_matrix, batch = batch, group = covariates)
```
```{r}
ensembl_ids <- rownames(adjusted)
ensembl_ids <- str_split_fixed(ensembl_ids, pattern = "\\.", n = 2)[,1]
info <- getBM(attributes=c("hgnc_symbol",
                           "ensembl_gene_id"),
                  filters = c("ensembl_gene_id"),
                  values = ensembl_ids,
                  mart = mart,
                  useCache=FALSE)
```
```{r}
genes_interest <- c("FAP", "ITGB1", "ACTA2", "PDPN", "PDGFRB")
info_unique <- info[!duplicated(info$hgnc_symbol),]
rownames(info_unique) <- info_unique$hgnc_symbol
genes_interest_info <- info_unique[genes_interest,] 
ids <- which(ensembl_ids %in% genes_interest_info$ensembl_gene_id)
adjusted_genes_interest <- adjusted[ids,]
colnames(adjusted_genes_interest) <- rownames(metadata_no_inhouse)
rownames(adjusted_genes_interest) <- genes_interest
```

```{r}
adjusted_genes_interest_exprs_t <- as.data.frame(t(adjusted_genes_interest))
adjusted_genes_interest_exprs_t$Subpopulation <- metadata_no_inhouse$Subtype
adjusted_genes_interest_exprs_t$Study <- metadata_no_inhouse$Study
adjusted_genes_interest_exprs_t$Tumor_JuxtaTumor <- metadata_no_inhouse$Tumor_JuxtaTumor
```

```{r}
plots_out <- lapply(FUN = caf_plot_function_study, X = genes_interest, input_table = adjusted_genes_interest_exprs_t)
ggar_obj <- ggarrange(plotlist = plots_out, common.legend = TRUE, legend = "right") # rel_heights values control title margins
title <- ggdraw() + draw_label("Batch-corrected expression of marker genes\naccording to CAF subpopulation", fontface='bold')
ggar_obj_annotated <- annotate_figure(ggar_obj, bottom = text_grob("CAF Subpopulation"), top = text_grob("VSD normalised expression of marker genes\naccording to CAF subpopulation", color = "black", face = "bold", size = 10))
ggar_obj_annotated

plots_out <- lapply(FUN = caf_plot_function_tumour_juxtatumour, X = genes_interest, input_table = adjusted_genes_interest_exprs_t)
ggar_obj <- ggarrange(plotlist = plots_out, common.legend = TRUE, legend = "right") # rel_heights values control title margins
title <- ggdraw() + draw_label("Batch-corrected expression of marker genes\naccording to CAF subpopulation", fontface='bold')
ggar_obj_annotated <- annotate_figure(ggar_obj, bottom = text_grob("CAF Subpopulation"), top = text_grob("VSD normalised expression of marker genes\naccording to CAF subpopulation", color = "black", face = "bold", size = 10))
ggar_obj_annotated
```

 -  Carry out batch correction between studies, cannot account for differences in subpopulations here
 -  Differential expression analysis CAF vs TAN
 -  We cannot include patient in the formula here as we don't have matched CAF TAN

```{r}
counts_matrix_all <- assay(dds)
batch_all <- metadata$Study
covariates_all <- metadata$Tumor_JuxtaTumor
  #Subpopulation = metadata_no_inhouse$Subtype)
                        
adjusted_all <- ComBat_seq(counts = counts_matrix_all, batch = batch_all, group = covariates_all)
```

```{r}
# have to divide everything by 10 because the max digit is ~10x the max machine integer
dds_adjusted_all <- DESeqDataSetFromMatrix(countData = round(adjusted_all/10), colData = metadata, design = ~ Study + Tumor_JuxtaTumor)
dds_adjusted_all$Tumor_JuxtaTumor <- relevel(dds_adjusted_all$Tumor_JuxtaTumor, ref = "juxtatumor")
dds_adjusted_all <- DESeq(dds_adjusted_all)
```

### Differentially Expressed Genes CAF vs TAN (Tumor vs Juxta-Tumor)

```{r Functions for extracting DE results}
get_upregulated <- function(df){

    key <- intersect(rownames(df)[which(df$log2FoldChange>=1)], rownames(df)[which(df$padj<=0.05)])

  results <- as.data.frame((df)[which(rownames(df) %in% key),])
    return(results)
}

get_downregulated <- function(df){

    key <- intersect(rownames(df)[which(df$log2FoldChange<=-1)], rownames(df)[which(df$padj<=0.05)])

    results <- as.data.frame((df)[which(rownames(df) %in% key),])
    return(results)
}

annotate_de_genes <- function(df, filter_by){
    # if your df has hgnc_symbol as rowmaes, filter by that, if it is the ENSG1234.12, use "ensembl_gene_id_version", if it is the regular engs, filter by "ensembl_gene_id"
    filter_by_string <- as.character(filter_by)
    df$gene_symbol <- rownames(df)
    colnames(df)[6] <- filter_by_string
    #print(df)
    mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host="uswest.ensembl.org")
    info <- getBM(attributes=c("hgnc_symbol",
                               "ensembl_gene_id_version",
                               "chromosome_name",
                               "start_position",
                               "end_position",
                               "strand",
                               "entrezgene_description"),
                  filters = c(filter_by_string),
                  values = df[,6],
                  mart = mart,
                  useCache=FALSE)

    tmp <- merge(df, info, by=filter_by_string)
    tmp$strand <- gsub("-1", "-", tmp$strand)
    tmp$strand <- gsub("1", "+", tmp$strand)
    #tmp$hgnc_symbol <- make.names(tmp$hgnc_symbol, unique = T)
    tmp <- tmp[!grepl("CHR", tmp$chromosome_name),]

    output_col <- c("Gene", "Ensembl ID", "Chromosome", "Start", "Stop", "Strand", "Description", "Log2FC", "P-value", "Adj P-value")
    tmp <- subset(tmp, select=c(hgnc_symbol, ensembl_gene_id_version, chromosome_name, start_position, end_position, strand, entrezgene_description, log2FoldChange, pvalue, padj))
    colnames(tmp) <- output_col

    if(min(tmp$Log2FC) > 0){
        tmp <- tmp[order(-tmp$Log2FC),]
    }else{
        tmp <- tmp[order(tmp$Log2FC),]
    }

    return(tmp)

}
```

```{R, message=F, warning=F}
dds <- dds_adjusted_all
res <- results(dds, filterFun = ihw, alpha = 0.05, c("Tumor_JuxtaTumor", "tumor", "juxtatumor"))
lfc <- lfcShrink(dds = dds, res = res, coef = 6, type = "apeglm")
lfc_df <- as.data.frame(lfc)

up <- get_upregulated(lfc_df)
down <- get_downregulated(lfc_df)

up <- annotate_de_genes(up, filter_by = "ensembl_gene_id_version")
down <- annotate_de_genes(down, filter_by = "ensembl_gene_id_version")
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
master_lfc1 <- rbind(up,down)
# want to create 'background' gene set entrez id + LFC values for all genes
info <- getBM(attributes = c("hgnc_symbol", "entrezgene_id", "ensembl_gene_id_version"),
              mart=mart, filters = "ensembl_gene_id_version", values = rownames(lfc_df))

lfc_df$Gene <- rownames(lfc_df)
tmp <- merge(info, lfc_df, by.x="ensembl_gene_id_version", by.y="Gene")

background <- tmp$log2FoldChange
names(background) <- tmp$hgnc_symbol
background <- sort(background, decreasing = TRUE)

# make sure you have all DEGs here (LFC > 1 and pval cutoff i.e maser_lfc1 genes)
degtmp <- subset(tmp, tmp$hgnc_symbol %in% master_lfc1$Gene)
deg <- degtmp$entrezgene_id
deg <- na.omit(deg)

#deg.df <- bitr(deg, fromType = "ENTREZID",
 #               toType = c("ENSEMBL", "SYMBOL"),
  #              OrgDb = org.Hs.eg.db)
```

#### Hallmarks GSEA

```{r}
hmarks <- read.gmt("~/Downloads/h.all.v7.4.symbols.gmt")

egmt <- GSEA(background,
             TERM2GENE=hmarks,
             pvalueCutoff = 0.1, 
             pAdjustMethod = 'BH',
             minGSSize = 1, 
             maxGSSize = 1000,
             by='fgsea',
             seed=11384191)

egmt_df <- egmt@result

egmt_subs <- subset(egmt_df, select=c(Description, enrichmentScore, NES, pvalue, p.adjust, core_enrichment ))

#DT::datatable(egmt_subs, rownames = FALSE, options=list(scrollX=T))
```

```{R, message=F, warning=F, fig.height=5, fig.width=7}
dotplot(egmt, title = "GSEA HALLMARKS", x="NES")
```

### Enrichment plots

```{R, message=F, warning=F, fig.width=10, fig.height=5}
library(enrichplot)
for(i in 1:5){
  p <- gseaplot2(egmt, geneSetID = i, title = egmt_df$Description[[i]], pvalue_table = T)
  plot(p)
}
```

#### GO Biological Processes GSEA

```{R,message=F, warning=F}
gobp <- read.gmt("~/Downloads/c5.go.bp.v7.4.symbols.gmt")

egmt <- GSEA(background,
             TERM2GENE=gobp,
             pvalueCutoff = 0.1, 
             pAdjustMethod = 'BH',
             minGSSize = 1, 
             maxGSSize = 1000,
             by='fgsea',
             seed=11384191)

egmt_df <- egmt@result
#egmt_df$Description <- gsub("GOBP_", "", egmt_df$Description)
#egmt_df$ID <- gsub("GOBP_", "", egmt_df$ID)

egmt_subs <- subset(egmt_df, select=c(Description, enrichmentScore, NES, pvalue, p.adjust, core_enrichment ))
egmt_subs[1:5,]
#DT::datatable(egmt_subs, rownames = FALSE, options=list(scrollX=T))
```

### Dot Plot

```{R, message=F, warning=F, fig.height=10, fig.width=12}
dotplot(egmt, title = "GO BIOLOGICAL PROCESSES", x="NES", showCategory=30, font.size = 8)
```

### Enrichment Plots

```{R, message=F, warning=F, fig.width=10, fig.height=5}
for(i in 1:5){
  p <- gseaplot2(egmt, geneSetID = i, title = egmt_df$Description[[i]], pvalue_table = T)
  plot(p)
}
```
### Data transformation

There are a number of options to choose from when normalising RNA-seq data, the main ones being:
 -  Take the log of the data and add a pseudocount.
 -  Variance stabilizing transformation (Anders and Huber 2010)
 -  Regularized logarithm transformation (Love, Huber, and Anders 2014)

The log+pseudocount approach tends to mean that lowly expressed genes have more of an effect. *Vst* and *rlog* bring these counts towards a central amount, making the data more homoskedastic. This allows them to be used in downstream processes which require homoskedastic data (e.g. PCA). The authors of DESeq2 recommend *vst* for large sample sizes such as ours as it is much faster than *rlog*


```{r}
# here, blind is FALSE as we don't want it to be blind to experimental design 
# recommended when transforming data for downstream analysis which will use the design information
vsd <- vst(dds, blind = FALSE)
```

```{r}
dds <- estimateSizeFactors(dds)

df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

There doesn't seem to much of a difference between the two methods of normalisation, only that the lowly expressed genes have been brought up to a minimum of ~4. 

### QC - PCA per study for outlier detection

```{r}
#studies <- levels(colData(vsd)$Study)
# write a function which takes in DESeq object & creates PCA of all studies
deseq_pca_studies <- function(dds_object){
  library(PCAtools)
  library(ggplot2)
  studies <- levels(colData(dds_object)$Study)
  pca_plots <- list()
  for (i in 1:length(studies)){
    study <- studies[i]
    samples <- colnames(dds_object)[which(colData(dds_object)$Study == study)]
    dds_object_study <- dds_object[,samples]
    stopifnot(dim(dds_object_study)[2] == length(samples))
    pca_plot_data <- plotPCA(dds_object_study, intgroup = c("Subtype", "Tumor_JuxtaTumor"), returnData = TRUE)
    percentVar <- round(100 * attr(pca_plot_data, "percentVar"))
    title <- paste("PCA plot of study:", study, sep = " ")
    #plot_labels <- rownames(pca_plot_data)
    #print(samples)
    #print(length(samples))
    pca_plot_data$samples <- rownames(pca_plot_data)
    pca_plot <- ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Subtype, shape = Tumor_JuxtaTumor, label = samples)) +
                geom_point(size =3) +
                geom_text(hjust="middle", vjust=1, data = subset(pca_plot_data, PC2 > 22)) +
                xlab(paste0("PC1: ", percentVar[1], "% variance")) +
                ylab(paste0("PC2: ", percentVar[2], "% variance")) +
                coord_fixed() +
                ggtitle(title)
    pca_plots[[i]] <- pca_plot
    #print(pca_plots[i])
  }
  #stopifnot(length(pca_plots) == length(studies))
  return(pca_plots)
}
```

```{r}
#output <- deseq_pca_studies(vsd)
#output
#output
```

```{r}
plotPCA(vsd, intgroup = c("Study", "Subtype"))

```


### Batch correction

The following graphs demonstrate the batch effects

```{r}
ann_colors = list(Study=c(EGAD00001004810 = "forestgreen",  EGAD00001003808 = "gold", EGAD00001006144 = "blue", EGAD00001005744 = "magenta3", InHouse = "black"),
                  Subtype = c(S1 = "dodgerblue4", S3 = "chartreuse1", S4 = "grey67", Unknown = "palevioletred"),
                  Tumor_JuxtaTumor = c(tumor = "royalblue", juxtatumor = "red")
                  )
```

```{r Generate distance matrix + generate heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
sampleDists <- dist(t(assay(vsd)))
# don't want files column to be on heatmap
subset_coldata <- subset(coldata, select = -c(files))
sampleDistMatrix <- as.matrix( sampleDists )
rownames(subset_coldata) <- subset_coldata$names
subset_coldata$names <- NULL

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

## Clinical Correlations

```{R, message=F, warning=F, fig.width=10}
metadata_pca <- metadata[,1:3]
vsd_matrix <- as.matrix(assay(vsd))
p <- pca(vsd_matrix, metadata = metadata_pca)

  eigencorplot(p,
    components = getComponents(p, 1:10),
    metavars = colnames(metadata_pca),
    col = c('white', 'cornsilk1', 'gold', 'forestgreen', 'darkgreen'),
    cexCorval = 0.7,
    colCorval = 'black',
    fontCorval = 2,
    posLab = 'bottomleft',
    rotLabX = 45,
    posColKey = 'top',
    cexLabColKey = 1.5,
    scale = TRUE,
    corFUN = 'pearson',
    corUSE = 'pairwise.complete.obs',
    corMultipleTestCorrection = 'none',
    main = 'PC1-10 clinical correlations',
    colFrame = 'white',
    plotRsquared = TRUE)
```

```{r}
plotPCA(vsd, intgroup = c("Study", "Subtype"))
```

```{r}
pcaData <- plotPCA(vsd, intgroup = c("Study", "Subtype"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData, aes(x = PC1, y = PC2, color = Study, shape = Subtype)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")
```

Next step is to look at batch correction.

#### Method 1: Limma's remove batch effect function

```{r message=FALSE}
library(limma)
```

```{r create copies of vsd to avoid confusion}
vsd_remove_batch_intercept <- vsd
vsd_remove_batch_tumor_juxta_subtype <- vsd
vsd_remove_batch_tumor_juxta_only <- vsd
```


```{r removeBatchEffect model matrix with only an intercept term}
mat <- assay(vsd_remove_batch_intercept)
# only intercept term in model matrix, don't include Subtype or Tumor_JuxtaTumor
mm <- model.matrix(~1, colData(vsd_remove_batch_intercept))
# remove study batch effect
mat <- limma::removeBatchEffect(mat, batch = vsd_remove_batch_intercept$Study, design = mm)
assay(vsd_remove_batch_intercept) <- mat
pca_batch_correct_intercept <- plotPCA(vsd, intgroup = c("Subtype","Study")) + 
  aes(x = PC1, y = PC2, color = colData(vsd_remove_batch_intercept)$Subtype, shape =colData(vsd_remove_batch_intercept)$Study ) +
  labs(color='Subpopulation', shape = "Study") +
  ggtitle("PCA of VST data batch corrected using \nlimma::removeBatchEffect with intercept term only")
pca_batch_correct_intercept
```

```{r removeBatchEffect model matrix with only an intercept term, heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
# same template as previous heatmap
sampleDists <- dist(t(assay(vsd_remove_batch_intercept)))
sampleDistMatrix <- as.matrix( sampleDists )

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```

Limma's removeBatchEffect function requires a design matrix as input, this is the "treatment conditions" we wish to preserve. It is usually the design matrix with all experimental factors other than batch effects. The ideal scenario would be to include Subtype in this matrix. However, this treats `Unknown` as its own subtype, and so will preserve differences between the InHouse samples and the other samples, as seen in the below PCA plot. This is contrary to what we want when assigning our samples to a cluster.

```{r removeBatchEffect tumor juxta + subtype model matrix}
mat <- assay(vsd_remove_batch_tumor_juxta_subtype)
# create model matrix, full model matrix with Tumor_JuxtaTumor and Subtype
mm <- model.matrix(~Tumor_JuxtaTumor+Subtype, colData(vsd_remove_batch_tumor_juxta_subtype))
mat <- limma::removeBatchEffect(mat, batch = vsd_remove_batch_tumor_juxta_subtype$Study, design = mm)
assay(vsd_remove_batch_tumor_juxta_subtype) <- mat

pca_batch_correct_tumor_juxta_subtype <- plotPCA(vsd_remove_batch_tumor_juxta_subtype, intgroup = c("Subtype","Study")) + 
  aes(x = PC1, y = PC2, color = colData(vsd_remove_batch_tumor_juxta_subtype)$Subtype, shape =colData(vsd_remove_batch_tumor_juxta_subtype)$Study ) +
  labs(color='Subpopulation', shape = "Study") +
  ggtitle("PCA of VST data batch corrected using \nlimma::removeBatchEffect with Tumor_JuxtaTumor\nand Subtype in model")
pca_batch_correct_tumor_juxta_subtype
```

```{r removeBatchEffect model matrix with Study+Subpopulation, heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
# same template as previous heatmap

sampleDists <- dist(t(assay(vsd_remove_batch_tumor_juxta_subtype)))
sampleDistMatrix <- as.matrix( sampleDists )

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```

We can see from the PCA and heatmap above that including condition (tumour-juxtatumour) and subtype in our model matrix leads to good separation of the subtypes, and a lack of clustering by batch. The problem with this approach is that it treats our `Unknown` subtype as its own subtype, which means it will cluster on its own no matter what we do. Is there a way around this? Frozen surrogate variable analysis could be an option, but this is usually used on microarray data, I don't know if it can be used on RNA-sequencing data.

```{r removeBatchEffect tumor juxta only matrix}
mat <- assay(vsd)
mm <- model.matrix(~Tumor_JuxtaTumor, colData(vsd))
mat_batch_removed_limma <- limma::removeBatchEffect(mat, batch = vsd$Study, design = mm)
```

In the PCA plot above, we have not told the `removeBatchEffect` function about our known subtypes, only whether the samples were taken from Tumor or Juxta-Tumor. It does not know to preserve differences between subpopulations when removing batch effects. Less variance being explained by PC1 than in the first scenario (39% vs 68%). In the first PCA plot, our in-house samples of unknown subtype cluster together on their own. 


```{r run BatchQC on our data}
#mat <- assay(vsd)
#modmatrix <- model.matrix(~as.factor(Tumor_JuxtaTumor), colData(vsd))
#batchQC(mat, batch=coldata$Study, condition= coldata$Subtype, 
      #  report_file = "batchqc_caf_data_not_corrected.html",
       # report_dir = ".", view_report = FALSE, interactive = FALSE
      #  )
```


```{r}
coldata_pca <- coldata
rownames(coldata_pca) <- coldata_pca$names
coldata_pca$names <- NULL
coldata_pca$files <- NULL
p <- pca(mat, metadata = coldata_pca)

```

```{r}
ggplot(p$rotated, aes(x = PC1, y = PC2, color = p$metadata$Study, shape = p$metadata$Subtype)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  labs(color='Study', shape = "Subpopulation") +
  ggtitle("PCA with transormed data after batch correction")
```

```{r Batch correction using Combat-Seq}
batch <- coldata$Study
mm <- model.matrix(~Tumor_JuxtaTumor, colData(vsd))
```


# Predicting subpopulation(s) present in In-House samples

## Assigning in-house samples to a CAF subtype using K-nearest neighbours

1. Split mat into our known (training) and unknown (testing) subpopulations
2. Run KNN


```{r create separate variables for known and unknown CAF subtypes}
mat_t <- t(mat)
mat_known <- mat_t[coldata$names[which(coldata$Subtype != "Unknown")],]
coldata_known <- coldata[coldata$Subtype != "Unknown",]
mat_unknown <- mat_t[coldata$names[which(coldata$Subtype == "Unknown")],]
coldata_unknown <- coldata[coldata$Subtype == "Unknown",]
```

```{r}
library(class)
```

```{r}
##create a random number equal 90% of total number of rows
 ran <- sample(1:nrow(mat_known),0.9 * nrow(mat_known))
 ##training dataset extracted
 mat_train <- mat_known[ran,]
 
 ##test dataset extracted
 mat_test <- mat_known[-ran,]
```

```{r}
# these are our labels
caf_target_category <- coldata_known[ran,4]
caf_test_category <- coldata_known[-ran,4]
```

```{r run KNN}
pr <- knn(mat_train,mat_test,cl=caf_target_category,k=21)
```

```{r}
 tab <- table(pr,caf_test_category)
tab
```
```{r}
accuracy <- function(x){sum(diag(x)/(sum(rowSums(x)))) * 100}
 accuracy(tab)
```
```{r}
  outputs <- c()
for (i in 1:50){
  pr <- knn(mat_train,mat_test,cl=caf_target_category,k=i)
  number <- i
  tab <- table(pr,caf_test_category)
  accuracy_out <- accuracy(tab)
  outputs <- c(outputs, number = accuracy_out)
}
```

```{r}
plot(outputs)
```

```{r}
  prediction <- knn(mat_train,mat_unknown,cl=caf_target_category,k=13)
names(prediction) <- rownames(mat_unknown)
prediction
```
They are all predicted to be S1 using this initial application of the algorithm except one sample.
Possibly use interquartile range to improve performance?

## Deconvolution using CIBERSORTx

CIBERSORTx [@Newman2019] is the most commonly used tool for cell-type deconvolution. It is a machine learning method which carries out batch correction, correcting for between-platform differences in expression data. In this case, the signature matrix was made using all of the available CAF subpopulation data (40 S1 samples, 25 S3 samples, 24 S4 samples). It is possible to infer the proportions of the different subpopulations as well as a subpopulation-specific gene expression matrix for each sample. Either scRNA-seq, bulk RNA-seq or microarray data can be used as the reference.

- Create signature matrix for CIBERSORTx
- TPM normalise data,probably optional
- Run CIBERSORTx to figure out proportions of S1, S3 and S4 in In-house samples

The files for CIBERSORT (mixture file, reference data and phenotype classes file), were prepared using the `cibersortx_prepare_files.R` script.

`CIBERSORTx` was run using the following command:
```
docker run -v /home/kevin/Documents/PhD/cibersort/caf_subpopulation/infiles:/src/data -v /home/kevin/Documents/PhD/cibersort/caf_subpopulation/outfiles:/src/outdir cibersortx/fractions --username k.ryan45@nuigalway.ie --token b7f03b943ade9b4146dc2126b4ac9d19 --single_cell FALSE --refsample caf_subtypes_tpm_for_sig_matrix.txt --mixture caf_tpm_mixture.txt --rmbatchBmode TRUE --outdir /home/kevin/Documents/PhD/cibersort/caf_subpopulation/outfiles --phenoclasses /home/kevin/Documents/PhD/cibersort/caf_subpopulation/infiles/phenoclasses_caf.txt
```

```{r CIBERSORT initial results}
cibersort_results <- read.table("/home/kevin/Documents/PhD/cibersort/caf_subpopulation/outfiles/CIBERSORTx_Adjusted.txt", header = T)
cibersort_results
```

Results look strange with P-value of 9999. CIBERSORTx was run using the Docker image and using the GUI, and different results were obtained. When the Docker image was used and the number of permutations was changed from 0 to 100, the p-value changed to 0.000. The proportions look different between the two methods, with the GUI predicting about 0.75 S1 with the rest being S4 for most samples.


## References
