---
title: "CAF Subpopulation Analysis"
author: "Kevin Ryan"
date: "`r Sys.time()`"
bibliography: citations.bib
output: 
  github_document:
     toc: true
     toc_depth: 3 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Cancer-associated fibroblasts (CAFs) are a heterogeneous cell type found in the tumour microenvironment. They have a wide array of functions, and tend to be immunosuppressive and cancer-promoting. There have been many attempts to characterise subpopulations of CAFs, with much transcriptomic analysis being carried out in the Mechta-Grigoriou lab in Institut Curie. They have identified 4 'subpopulations' which can be separated based on the expression of different markers:

 -  S1: FAP^High^, CD29^Med-High^, α^SMAHigh^, PDPN^High^, PDGFRβ^High^
 -  S2: FAP^Neg^, CD29^Low^, αSMANeg-^Low^, PDPN^Low^, PDGFRβ^Low^
 -  S3: FAP^Neg-Low^, CD29^Med^, αSMA^Neg-Low^, PDPN^Low^, PDGFRβ^Low-Med^
 -  S4: FAP^Low-Med^, CD29^High^, αSMA^High^, PDPN^Low^, PDGFRβ^Med^

[@Pelon2020]

FACS gating strategies can be used to isolate these various subpopulations. The Mechta-Grigoriou group have done this and have generated bulk RNA-sequencing data for the S1, S3 and S4 subpopulations. They generated scRNA-sequencing data for the S1 subpopulation. This data was deposited on the European Genome Phenome Archive, and was accessed via a Data Transfer Agreement.

The following summarises the data obtained:

+---------------+---------------+-----------------------+-------------------------------+
| Subpopulation | Total samples | Studies (Samples)     |  Notes                        |
+===============+===============+=======================+===============================+
| S1            |  28           | - EGAD00001003808 (16)| -  3808 has 12xJuxta-tumor    |
|               |               | - EGAD00001005744 (5) | -  5744 5 samples from LN     |
|               |               | - EGAD00001006144 (7) | -  Sorting vs spreading       |
+---------------+---------------+-----------------------+-------------------------------+
| S2            | 0             | N/A                   | N/A                           |
|               |               |                       |                               |
+---------------+---------------+-----------------------+-------------------------------+
| S3            | 14            | - EGAD00001004810 (14)| -  4810 has 11xJuxta-tumor    |
|               |               |                       | -  Ovarian                    |
+---------------+---------------+-----------------------+-------------------------------+
| S4            | 15            | - EGAD00001003808 (10)| -  3808 has 9xJuxta-tumor     |
|               |               | - EGAD00001005744 (5) | -  5744 5 samples from LN     |
+---------------+---------------+-----------------------+-------------------------------+

With the juxta-tumour data, they got tumour and juxta-tumour samples from the same patient. However, I have not been able to figure out whether they came from the same patient. We could possibly use Optitype to determine HLA allele - match tumour and juxta tumour.

We also have scRNA-seq data for S1, labelled with 8 subpopulations of S1 CAFs. It may be possible to 

It is likely that sorting the cells using FACS alters the transcriptional properties of the cells compared to if they are separated using spreading approaches, as is seen in study `EGAD00001006144`. This is something that we will have to keep in mind.

The data was processed using nf-core/rnaseq version `3.8.1` using the default parameters. STAR/Salmon were used for alignment/quantification.

We would expect our tumour-associated normal to be most like the S3 subpopulation (usually accumulate in juxta-tumours). The S2 subpopulation has been found to accumulate more in luminal A breast cancer, whereas the S4 subpopulation tends to be present in Her2+ breast cancers. Unfortunately, data is not available for the S2 subpopulation and 11 of the 12 cancers encountered in our samples are Luminal A.

Combining RNA-sequencing datasets from different studies can be very challenging. We can expect batch effects to be present, so it might not be possible to determine whether differences we observe are due to actual biological effects or technical artifacts. In addition, a recent study suggests that DESeq2 and edgeR (the most popular differential expression tools) experience large rates of false positives when used with large sample sizes [@Li2022]. However, this assertion has been refuted, and it has been suggested that the Li 2022 study did not apply appropriate batch correction and quality control ([Twitter thread](https://threadreaderapp.com/thread/1513468597288452097.html) from Mike Love and associated [code on GitHub](https://github.com/mikelove/preNivolumabOnNivolumab/blob/main/preNivolumabOnNivolumab.knit.md)). One of the datasets (`EGAD00001006144`) was produced using stranded RNA-seq, whereas the other datasets were unstranded. This can lead to a lack of comparability of the datasets [@Zhao2020]. It may be necessary to drop this dataset from the analysis. All samples were prepared by poly(A) selection (use of oligo-dT).

# Analysis
## Preparation

Columns will be: Sample, Study, Subpopulation, Tumor_Juxtatumor

*Here we will be combining data from 5 studies. To begin with, we will only include the metadata available for all studies (except for our unknown CAF Subpopulation label). Breast cancer subtype is only available for certain studies and so is not included at this stage.*

There are also: ovarian cancer samples, EPCAM+ cells (an epithelial marker) and samples from lymph nodes. For the time being, I will not consider them.

```{r load packages, include = FALSE}
library(dplyr)
library(stringr)
library(biomaRt)
library(tximport)
library(DT)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(PCAtools)
library(dplyr)
library(SummarizedExperiment)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(glmpca)
library(hexbin)
library(IHW)
library(sva)
library(ggpubr)
library(vsn)
httr::set_config(httr::config(ssl_verifypeer = FALSE))
library(tximeta)
library(BiocParallel)
library(ashr)
library(GSVA)
library(clusterProfiler)
library(org.Hs.eg.db)
library(HGNChelper)
```

```{r read in metadata, include=FALSE}
EGAD_4810 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001004810/delimited_maps/Run_Sample_meta_info.map", 
                        sep = ";")
EGAD_4810_cancers <- str_split_fixed(EGAD_4810$V1, pattern = "=", n = 2)[,2]
EGAD_4810_keep <- which(EGAD_4810_cancers == "BC")
EGAD_4810_filtered <- EGAD_4810[EGAD_4810_keep,]
EGAD_4810_meta <- data.frame(
  Sample = str_split_fixed(EGAD_4810_filtered$V4, pattern = "=", n = 2)[,2],
  Study = "EGAD00001004810",
  Subpopulation = "S3",
  Tumor_JuxtaTumor = tolower(str_split_fixed(EGAD_4810_filtered$V3, pattern = " ", n = 2)[,2]),
  Strandedness = "unstranded",
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001004810_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_3808 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001003808/meta_CAF-S1_S4_BC_47samples.txt", 
                        header = T, sep = "\t")
EGAD_3808_meta <- data.frame(
  Sample = EGAD_3808$Sample.Name,
  Study = "EGAD00001003808",
  Subpopulation = EGAD_3808$subset,
  Tumor_JuxtaTumor = EGAD_3808$Type,
  Strandedness = "unstranded",
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001003808_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_6144 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001006144/meta_7samples.txt", 
                        header = T,sep = "\t")
EGAD_6144_meta <- data.frame(
  Sample = paste("CAF_Culture_", EGAD_6144$Sample.Name, sep = ""),
  Study = "EGAD00001006144",
  Subpopulation = "S1",
  Tumor_JuxtaTumor = "tumor",
  Strandedness = "reverse",
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001006144_nfcore_results/star_salmon",
  row.names = 1
)
EGAD_5744 <- read.table("/home/kevin/Documents/PhD/CAF_data/EGAD00001005744/metaData_Pelon_et_al.txt", 
                        header =T, check.names = F)
EGAD_5744$Sample.Name <- gsub(pattern = "\\.", replacement = "-", x = EGAD_5744$Sample.Name )
EGAD_5744_filtered <- EGAD_5744[!(EGAD_5744$subset == "EPCAM+") & (EGAD_5744$Type == "T"),]
EGAD_5744_meta <- data.frame(
  Sample = EGAD_5744_filtered$Sample.Name,
  Study = "EGAD00001005744",
  Subpopulation = EGAD_5744_filtered$subset,
  Tumor_JuxtaTumor = "tumor",
  Strandedness = "unstranded",
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/EGAD00001005744_nfcore_results/star_salmon",
  row.names = 1
)
barkley_samples <- read.csv("/home/kevin/Documents/PhD/rna_seq_bc/metadata/reformat_samples.csv", 
                            header = T, row.names = "samples", check.names = F)
barkley_samples_meta <- data.frame(
  Sample = row.names(barkley_samples),
  Study = "InHouse",
  Subpopulation = "Unknown",
  Tumor_JuxtaTumor = ifelse(barkley_samples$Condition == "Tumour", "tumor", "juxtatumor"),
  Strandedness = "unstranded",
  directory = "/home/kevin/Documents/PhD/CAF_data/nfcore_results/inhouse_data_nfcore_results_version_3_8_1/star_salmon",
  row.names = 1
)
metadata <- rbind.data.frame(EGAD_4810_meta, EGAD_3808_meta, 
                             EGAD_6144_meta, EGAD_5744_meta, barkley_samples_meta)
metadata$Tumor_JuxtaTumor <- gsub(x = metadata$Tumor_JuxtaTumor, pattern = "-", replacement = "")
#write.table(metadata, file = "/home/kevin/Documents/PhD/subtypes/caf-subtype-analysis/metadata_caf_subtypes.txt", quote = F, sep = "\t", row.names = T)
metadata_no_inhouse <- rbind.data.frame(EGAD_4810_meta, EGAD_3808_meta, 
                             EGAD_6144_meta, EGAD_5744_meta)
metadata_no_inhouse$Tumor_JuxtaTumor <- gsub(x = metadata_no_inhouse$Tumor_JuxtaTumor, pattern = "-", replacement = "")

metadata_no_inhouse_no6144 <- rbind.data.frame(EGAD_4810_meta, EGAD_3808_meta, EGAD_5744_meta)
```

## Read in data

Samples were processed with nf-core/rnaseq version `3.8.1`
Salmon was used in alignment mode so there is no salmon index, so there is no checksum to import the metadata. Therefore, the parameters recommended in the [tximeta vignette](https://bioconductor.org/packages/release/bioc/vignettes/tximeta/inst/doc/tximeta.html#What_if_checksum_isn%E2%80%99t_known) were used to summarise transcript counts to the gene level, using a tx2gene file constructed using `generate_tx2gene_table.R`.

```{r}
files <- file.path(metadata$directory, rownames(metadata), "quant.sf")
coldata <- data.frame(files, names=rownames(metadata), Study = metadata$Study, 
                      Subpopulation = metadata$Subpopulation, 
                      Tumor_JuxtaTumor = metadata$Tumor_JuxtaTumor,
                      stringsAsFactors=FALSE)
# tx2gene file for the gencode v31 file used in the analysis, generated using generate_tx2gene_table.R
#tx2gene <- read_tsv("/home/kevin/Documents/PhD/references/tx2gene_gencode_v31.txt")

# tx2gene but using the hgnc symbol instead of ensembl gene id version
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", #host="https://www.ensembl.org")
                host="uswest.ensembl.org")
tx2gene <- getBM(attributes = c("ensembl_transcript_id_version", "hgnc_symbol"), mart = mart, useCache = FALSE)

# salmon was used in alignment mode so there is no salmon index, therefore there is no checksum to import the metadata 
# txOut = FALSE means to summarise to gene level (i.e. don't give out transcripts, give out gene level)
se <- tximeta(coldata, skipMeta=TRUE, txOut=FALSE, tx2gene=tx2gene)
#se_hgnc <- tximeta(coldata, skipMeta=TRUE, txOut=FALSE, tx2gene=tx2gene_hgnc)
# read in using tx
```

```{r Carry out all steps up to vsd step for se_hgnc}
dds <- DESeqDataSet(se, design = ~1)

# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds)) >= 10
# only keep rows (genes) for which keep is TRUE
dds <- dds[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds) >= 10) >= X
dds <- dds[keep,]
vsd <- vst(dds, blind = FALSE)
```


```{r read in data without inhouse data}
files_no_inhouse <- file.path(metadata_no_inhouse$directory, rownames(metadata_no_inhouse), "quant.sf")
coldata_no_inhouse <- data.frame(files = files_no_inhouse, names=rownames(metadata_no_inhouse), Study = metadata_no_inhouse$Study, 
                      Subpopulation = metadata_no_inhouse$Subpopulation, 
                      Tumor_JuxtaTumor = metadata_no_inhouse$Tumor_JuxtaTumor,
                      Strandedness = metadata_no_inhouse$Strandedness,
                      stringsAsFactors=FALSE)
# tx2gene file for the gencode v31 file used in the analysis
#tx2gene <- read_tsv("/home/kevin/Documents/PhD/references/tx2gene_gencode_v31.txt")
se_no_inhouse <- tximeta(coldata = coldata_no_inhouse, skipMeta=TRUE, txOut=FALSE, tx2gene=tx2gene)
```
```{r}
idx <- which(se_no_inhouse$Study != "EGAD00001006144")
se_no_inhouse_no_6144 <- se_no_inhouse[,idx]
```


```{r Apply common filters according to DESeq2 vignette to remove lowly expressed genes}
# do no design for the time being, Subpopulation + Batch gives error - model matrix not full rank
# this function stores input values, intermediate calculations and results of DE analysis - makes counts non-negative integers
dds_no_inhouse <- DESeqDataSet(se_no_inhouse, design = ~1)

# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds_no_inhouse)) >= 10
# only keep rows (genes) for which keep is TRUE
dds_no_inhouse <- dds_no_inhouse[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds_no_inhouse) >= 10) >= X
dds_no_inhouse <- dds_no_inhouse[keep,]
dds_no_inhouse_copy <- dds_no_inhouse
ntd <- normTransform(dds_no_inhouse)
```

```{r}
# do no design for the time being, Subpopulation + Batch gives error - model matrix not full rank
# this function stores input values, intermediate calculations and results of DE analysis - makes counts non-negative integers
dds <- DESeqDataSet(se, design = ~1)
# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds)) >= 10
# only keep rows (genes) for which keep is TRUE
dds <- dds[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds) >= 10) >= X
dds <- dds[keep,]
ntd <- normTransform(dds)
```

```{r}
dds_noinhouse_no6144 <- DESeqDataSet(se_no_inhouse_no_6144, design = ~1)
# returns a vector of whether the total count of each gene is >= 10 (True or false)
keep <- rowSums(counts(dds_noinhouse_no6144)) >= 10
# only keep rows (genes) for which keep is TRUE
dds_noinhouse_no6144 <- dds_noinhouse_no6144[keep,]
# at least X samples with a count of 10 or more, where X can be chosen as the sample size of the smallest group of samples
X <- 7
keep <- rowSums(counts(dds_noinhouse_no6144) >= 10) >= X
dds_noinhouse_no6144 <- dds_noinhouse_no6144[keep,]
```

### Data transformation

There are a number of options to choose from when normalising RNA-seq data, the main ones being:
 *  Take the log of the data and add a pseudocount.
 *  Variance stabilizing transformation (Anders and Huber 2010)
 *  Regularized logarithm transformation (Love, Huber, and Anders 2014)

The log+pseudocount approach tends to mean that lowly expressed genes have more of an effect. *Vst* and *rlog* bring these counts towards a central amount, making the data more homoskedastic. This allows them to be used in downstream processes which require homoskedastic data (e.g. PCA). The authors of DESeq2 recommend *vst* for large sample sizes such as ours as it is much faster than *rlog*.

```{r Mean variance relationship using log(x+1) transformation}
meanSdPlot(assay(ntd))
```

```{r Mean variance relationship using variance stabilising transformation}
# blind is FALSE when we don't want it to be blind to experimental design 
# recommended when transforming data for downstream analysis which will use the design information
# blind TRUE is recommended when transforming for QA purposes
vsd <- vst(dds, blind = TRUE)
meanSdPlot(assay(vsd))
```


```{r}
dds <- estimateSizeFactors(dds)

df <- bind_rows(
  as_data_frame(log2(counts(dds, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)  
```

There doesn't seem to much of a difference between the two methods of normalisation, only that the lowly expressed genes have been brought up to a minimum of ~4. 

## QC - PCA per study

Function not working at present

```{r}
#studies <- levels(colData(vsd)$Study)
# write a function which takes in DESeq object & creates PCA of all studies
deseq_pca_studies <- function(dds_object){
  library(PCAtools)
  library(ggplot2)
  studies <- levels(colData(dds_object)$Study)
  pca_plots <- list()
  for (i in 1:length(studies)){
    study <- studies[i]
    samples <- colnames(dds_object)[which(colData(dds_object)$Study == study)]
    dds_object_study <- dds_object[,samples]
    stopifnot(dim(dds_object_study)[2] == length(samples))
    pca_plot_data <- plotPCA(dds_object_study, intgroup = c("Subpopulation", "Tumor_JuxtaTumor"), returnData = TRUE)
    percentVar <- round(100 * attr(pca_plot_data, "percentVar"))
    title <- paste("PCA plot of study:", study, sep = " ")
    #plot_labels <- rownames(pca_plot_data)
    #print(samples)
    #print(length(samples))
    pca_plot_data$samples <- rownames(pca_plot_data)
    pca_plot <- ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Subpopulation, shape = Tumor_JuxtaTumor, label = samples)) +
                geom_point(size =3) +
                geom_text(hjust="middle", vjust=1, data = subset(pca_plot_data, PC2 > 22)) +
                xlab(paste0("PC1: ", percentVar[1], "% variance")) +
                ylab(paste0("PC2: ", percentVar[2], "% variance")) +
                coord_fixed() +
                ggtitle(title)
    pca_plots[[i]] <- pca_plot
    #print(pca_plots[i])
  }
  #stopifnot(length(pca_plots) == length(studies))
  return(pca_plots)
}
```

```{r}
#output <- deseq_pca_studies(vsd)
#output
#output
```

```{r}
plotPCA(vsd, intgroup = c("Study", "Subpopulation"))
```

## Batch Correction

### Clinical Correlations without Batch Correction

PCA plots as per https://github.com/kevinblighe/PCAtools


```{r message = FALSE, warning = FALSE, fig.height = 10, fig.width = 15, fig.cap = 'Merged panel of all PCAtools plots before batch correction'}
vsd_mat <- assay(vsd)
#keep <- rownames(vsd_mat)[!duplicated(rownames(vsd_mat))]
#vsd_mat <- vsd_mat[keep,]
metadata_pca <- metadata[,1:4]
p <- pca(vsd_mat, metadata = metadata_pca)
pscree <- screeplot(p, components = getComponents(p, 1:30),
    hline = 80, vline = 24, axisLabSize = 14, titleLabSize = 20,
    returnPlot = FALSE) +
    geom_label(aes(20, 80, label = '80% explained variation', vjust = -1, size = 8))
ppairs <- pairsplot(p, components = getComponents(p, c(1:3)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.8, gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'Tumor_JuxtaTumor',
    title = '', plotaxes = FALSE,
    margingaps = unit(c(0.01, 0.01, 0.01, 0.01), 'cm'),
    returnPlot = FALSE)
 pbiplot <- biplot(p,
    # loadings parameters
      showLoadings = TRUE,
      lengthLoadingsArrowsFactor = 1.5,
      sizeLoadingsNames = 4,
      colLoadingsNames = 'red4',
    # other parameters
      lab = NULL,
      colby = 'Tumor_JuxtaTumor', colkey = c('tumor'='royalblue', 'juxtatumor'='red3'),
      hline = 0, vline = c(-25, 0, 25),
      vlineType = c('dotdash', 'solid', 'dashed'),
      gridlines.major = FALSE, gridlines.minor = FALSE,
      pointSize = 5,
      legendPosition = 'none', legendLabSize = 16, legendIconSize = 8.0,
      #shape = 'Study', shapekey = c('Grade 1'=15, 'Grade 2'=17, 'Grade 3'=8),
      drawConnectors = FALSE,
      title = 'PCA bi-plot',
      subtitle = 'PC1 versus PC2',
      caption = '24 PCs ≈ 80%',
      returnPlot = FALSE)
 
 ploadings <- plotloadings(p, rangeRetain = 0.01, labSize = 3,
    title = 'Loadings plot', axisLabSize = 12,
    subtitle = 'PC1, PC2, PC3, PC4, PC5',
    caption = 'Top 1% variables',
    shape = 24, shapeSizeRange = c(4, 8),
    col = c('limegreen', 'black', 'red3'),
    legendPosition = 'none',
    drawConnectors = FALSE,
    returnPlot = FALSE)
 
 #peigencor <- #eigencorplot(p,
#    components = getComponents(p, 1:10),
#    metavars = c('Study','Subpopulation','Tumor_JuxtaTumor'),
#    cexCorval = 1.0,
#    fontCorval = 2,
#    posLab = 'all', 
#    rotLabX = 45,
#    scale = TRUE,
#    main = "PC clinical correlates",
#    cexMain = 1.5,
#    plotRsquared = FALSE,
#    corFUN = 'pearson',
#    corUSE = 'pairwise.complete.obs',
#    signifSymbols = c('****', '***', '**', '*', ''),
#    signifCutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
#    returnPlot = FALSE)
peigencor <- eigencorplot(p,
    components = getComponents(p, 1:10),
    metavars = colnames(metadata_pca),
    col = c('white', 'cornsilk1', 'gold', 'forestgreen', 'darkgreen'),
    cexCorval = 0.7,
    colCorval = 'black',
    fontCorval = 2,
    posLab = 'bottomleft',
    rotLabX = 45,
    posColKey = 'top',
    cexLabColKey = 1.5,
    scale = TRUE,
    corFUN = 'pearson',
    corUSE = 'pairwise.complete.obs',
    corMultipleTestCorrection = 'none',
    main = 'PC1-10 clinical correlations',
    colFrame = 'white',
    plotRsquared = TRUE)
 
  library(cowplot)
  library(ggplotify)

    top_row <- plot_grid(pscree, ppairs, pbiplot,
      ncol = 3,
      labels = c('A', 'B  Pairs plot', 'C'),
      label_fontfamily = 'serif',
      label_fontface = 'bold',
      label_size = 22,
      align = 'h',
      rel_widths = c(1.10, 0.80, 1.10))

    bottom_row <- plot_grid(ploadings,
      as.grob(peigencor),
      ncol = 2,
      labels = c('D', 'E'),
      label_fontfamily = 'serif',
      label_fontface = 'bold',
      label_size = 22,
      align = 'h',
      rel_widths = c(0.8, 1.2))

    plot_grid(top_row, bottom_row, ncol = 1,
      rel_heights = c(1.1, 0.9))
```

### Batch Correction with Combat-Seq

Here, batch correction is carried out using Study as our batch and Tumor_JuxtaTumor status as our group. This should remove differences between batches without removing differences between Tumor and JuxtaTumor. However, it does not preserve differences between CAF subpopulations. This will allow us to carry out DE analysis on these samples. Another method, which is recommended when carrying out differential expression analysis, is to include batch as a covariate in the model matrix, i.e. ~ Tumor_JuxtaTumor + Study (possibly???). However, this results in a model matrix of less than full rank, and so is not an option.

```{r Batch correction with ComBat_seq}
counts_matrix_all <- assay(dds)
batch_all <- metadata$Study
covariates_all <- metadata$Tumor_JuxtaTumor
adjusted_all <- ComBat_seq(counts = counts_matrix_all, batch = batch_all, group = covariates_all)
```

```{r Function to round numeric columns of dataframe to desired number of digits}
round_df <- function(x, digits) {
    # round all numeric variables
    # x: data frame 
    # digits: number of digits to round
    numeric_columns <- sapply(x, mode) == 'numeric'
    x[numeric_columns] <-  round(x[numeric_columns], digits)
    x
}
```

```{r VST transform batch corrected data + round}
adjusted_all_reduced <- adjusted_all/100
adjusted_all_reduced <- round_df(adjusted_all_reduced)
dds_batch_corrected <- DESeqDataSetFromMatrix(adjusted_all_reduced, colData = metadata, design = ~1)
vsd <- vst(dds_batch_corrected, blind = FALSE)
```


### Clinical Correlations after batch correction


```{r message = FALSE, warning = FALSE, fig.height = 10, fig.width = 15, fig.cap= 'Merged panel of all PCAtools plots after batch correction'}
p <- pca(assay(vsd), metadata = metadata_pca)
pscree <- screeplot(p, components = getComponents(p, 1:40),
    hline = 80, vline = 40, axisLabSize = 14, titleLabSize = 20,
    returnPlot = FALSE) +
    geom_label(aes(20, 80, label = '80% explained variation', vjust = -1, size = 8))
ppairs <- pairsplot(p, components = getComponents(p, c(1:3)),
    triangle = TRUE, trianglelabSize = 12,
    hline = 0, vline = 0,
    pointSize = 0.8, gridlines.major = FALSE, gridlines.minor = FALSE,
    colby = 'Tumor_JuxtaTumor',
    title = '', plotaxes = FALSE,
    margingaps = unit(c(0.01, 0.01, 0.01, 0.01), 'cm'),
    returnPlot = FALSE)
 pbiplot <- biplot(p,
    # loadings parameters
      showLoadings = TRUE,
      lengthLoadingsArrowsFactor = 1.5,
      sizeLoadingsNames = 4,
      colLoadingsNames = 'red4',
    # other parameters
      lab = NULL,
      colby = 'Tumor_JuxtaTumor', colkey = c('tumor'='royalblue', 'juxtatumor'='red3'),
      hline = 0, vline = c(-25, 0, 25),
      vlineType = c('dotdash', 'solid', 'dashed'),
      gridlines.major = FALSE, gridlines.minor = FALSE,
      pointSize = 5,
      legendPosition = 'none', legendLabSize = 16, legendIconSize = 8.0,
      #shape = 'Study', shapekey = c('Grade 1'=15, 'Grade 2'=17, 'Grade 3'=8),
      drawConnectors = FALSE,
      title = 'PCA bi-plot',
      subtitle = 'PC1 versus PC2',
      caption = '40 PCs ≈ 80%',
      returnPlot = FALSE)
 
 ploadings <- plotloadings(p, rangeRetain = 0.01, labSize = 4,
    title = 'Loadings plot', axisLabSize = 12,
    subtitle = 'PC1, PC2, PC3, PC4, PC5',
    caption = 'Top 1% variables',
    shape = 24, shapeSizeRange = c(4, 8),
    col = c('limegreen', 'black', 'red3'),
    legendPosition = 'none',
    drawConnectors = FALSE,
    returnPlot = FALSE)
 
 #peigencor <- #eigencorplot(p,
#    components = getComponents(p, 1:10),
#    metavars = c('Study','Subpopulation','Tumor_JuxtaTumor'),
#    cexCorval = 1.0,
#    fontCorval = 2,
#    posLab = 'all', 
#    rotLabX = 45,
#    scale = TRUE,
#    main = "PC clinical correlates",
#    cexMain = 1.5,
#    plotRsquared = FALSE,
#    corFUN = 'pearson',
#    corUSE = 'pairwise.complete.obs',
#    signifSymbols = c('****', '***', '**', '*', ''),
#    signifCutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 1),
#    returnPlot = FALSE)
peigencor <- eigencorplot(p,
    components = getComponents(p, 1:10),
    metavars = colnames(metadata_pca),
    col = c('white', 'cornsilk1', 'gold', 'forestgreen', 'darkgreen'),
    cexCorval = 0.7,
    colCorval = 'black',
    fontCorval = 2,
    posLab = 'bottomleft',
    rotLabX = 45,
    posColKey = 'top',
    cexLabColKey = 1.5,
    scale = TRUE,
    corFUN = 'pearson',
    corUSE = 'pairwise.complete.obs',
    corMultipleTestCorrection = 'none',
    main = 'PC1-10 clinical correlations',
    colFrame = 'white',
    plotRsquared = TRUE)

    top_row <- plot_grid(pscree, ppairs, pbiplot,
      ncol = 3,
      labels = c('A', 'B  Pairs plot', 'C'),
      label_fontfamily = 'serif',
      label_fontface = 'bold',
      label_size = 22,
      align = 'h',
      rel_widths = c(1.10, 0.80, 1.10))

    bottom_row <- plot_grid(ploadings,
      as.grob(peigencor),
      ncol = 2,
      labels = c('D', 'E'),
      label_fontfamily = 'serif',
      label_fontface = 'bold',
      label_size = 22,
      align = 'h',
      rel_widths = c(0.8, 1.2))

    plot_grid(top_row, bottom_row, ncol = 1,
      rel_heights = c(1.1, 0.9))
```


### Differential Expression Analysis CAF vs TAN

```{r}
#dds <- DESeqDataSet(countData = round(adjusted_all/10), colData = metadata, design = ~ Study + Tumor_JuxtaTumor)
```

### Surrogate variable analysis

#### svaseq with study EGAD00001006144

Looking for hidden sources of variation in the data. Results very difficult to interpret.

```{r}
dds_no_inhouse <- DESeq(dds_no_inhouse)
dat  <- counts(dds_no_inhouse, normalized = TRUE)
idx  <- rowMeans(dat) > 1
dat  <- dat[idx, ]
# we have Subpopulation and tumour-juxtatomour in the model matrix
mod  <- model.matrix(~ Subpopulation + Tumor_JuxtaTumor, colData(dds_no_inhouse))
mod0 <- model.matrix(~   1, colData(dds_no_inhouse))
svseq <- svaseq(dat, mod, mod0)
```

```{r}
par(mfrow = c(4, 5), mar = c(3,5,3,1))
for (i in 1:20) {
  stripchart(svseq$sv[, i] ~ dds_no_inhouse$Study, vertical = TRUE, main = paste0("SV", i))
  abline(h = 0)
 }
```


```{r}
#svseq$sv
par(mfrow = c(4, 5), mar = c(3,5,3,1))
for (i in 1:20) {
  stripchart(svseq$sv[, i] ~ dds_no_inhouse$Subpopulation, vertical = TRUE, main = paste0("SV", i))
  abline(h = 0)
 }
```

```{r}
#svseq$sv
par(mfrow = c(4, 5), mar = c(3,5,3,1))
for (i in 1:20) {
  stripchart(svseq$sv[, i] ~ dds_no_inhouse$Strandedness, vertical = TRUE, main = paste0("SV", i))
  abline(h = 0)
 }
```

#### svaseq without study EGAD00001006144

```{r}
dds_noinhouse_no6144 <- DESeq(dds_noinhouse_no6144)
dat  <- counts(dds_noinhouse_no6144, normalized = TRUE)
idx  <- rowMeans(dat) > 1
dat  <- dat[idx, ]
# we have Subpopulation and tumour-juxtatomour in the model matrix
mod  <- model.matrix(~ Subpopulation + Tumor_JuxtaTumor, colData(dds_noinhouse_no6144))
mod0 <- model.matrix(~   1, colData(dds_noinhouse_no6144))
svseq_no6144 <- svaseq(dat, mod, mod0)
```
```{r}
par(mfrow = c(4, 5), mar = c(3,5,3,1))
for (i in 1:ncol(svseq_no6144$sv)) {
  stripchart(svseq_no6144$sv[, i] ~ dds_noinhouse_no6144$Study, vertical = TRUE, main = paste0("SV", i))
  abline(h = 0)
 }
```


```{r}
#svseq$sv
par(mfrow = c(4, 5), mar = c(3,5,3,1))
for (i in 1:ncol(svseq_no6144$sv)) {
  stripchart(svseq_no6144$sv[, i] ~ dds_noinhouse_no6144$Subpopulation, vertical = TRUE, main = paste0("SV", i))
  abline(h = 0)
 }
```


### Differential expression analysis without the inhouse data and without study 6144

```{r Functions for extracting DE results}
get_upregulated <- function(df){

    key <- intersect(rownames(df)[which(df$log2FoldChange>=2)], rownames(df)[which(df$padj<=0.05)])

  results <- as.data.frame((df)[which(rownames(df) %in% key),])
    return(results)
}

get_downregulated <- function(df){

    key <- intersect(rownames(df)[which(df$log2FoldChange<=-2)], rownames(df)[which(df$padj<=0.05)])

    results <- as.data.frame((df)[which(rownames(df) %in% key),])
    return(results)
}

annotate_de_genes <- function(df, filter_by){
    # if your df has hgnc_symbol as rownames, filter by that, if it is the ENSG1234.12, use "ensembl_gene_id_version", if it is the regular engs, filter by "ensembl_gene_id"
    filter_by_string <- as.character(filter_by)
    df$gene_symbol <- rownames(df)
    colnames(df)[6] <- filter_by_string
    #print(df)
    mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl",  #host="https://www.ensembl.org")
                    host="uswest.ensembl.org")
    info <- getBM(attributes=c("hgnc_symbol",
                               "ensembl_gene_id_version",
                               "chromosome_name",
                               "start_position",
                               "end_position",
                               "strand",
                               "entrezgene_description",
                               "entrezgene_id"),
                  filters = c(filter_by_string),
                  values = df[,6],
                  mart = mart,
                  useCache=FALSE)

    tmp <- merge(df, info, by=filter_by_string)
    tmp$strand <- gsub("-1", "-", tmp$strand)
    tmp$strand <- gsub("1", "+", tmp$strand)
    #tmp$hgnc_symbol <- make.names(tmp$hgnc_symbol, unique = T)
    tmp <- tmp[!grepl("CHR", tmp$chromosome_name),]

    output_col <- c("Gene", "Ensembl ID", "Chromosome", "Start", "Stop", "Strand", "Description", "Log2FC", "P-value", "Adj P-value", "Entrez ID")
    tmp <- subset(tmp, select=c(hgnc_symbol, ensembl_gene_id_version, chromosome_name, start_position, end_position, strand, entrezgene_description, log2FoldChange, pvalue, padj, entrezgene_id))
    colnames(tmp) <- output_col

    if(min(tmp$Log2FC) > 0){
        tmp <- tmp[order(-tmp$Log2FC),]
    }else{
        tmp <- tmp[order(tmp$Log2FC),]
    }

    return(tmp)

}

anti_join_rownames <- function(df1, df2){
  anti_join((df1 %>% mutate(Symbol = rownames(df1))),
           (df2 %>% mutate(Symbol = rownames(df2))),
          by = 'Symbol')
}

filter_dfs_antijoin_rownames <- function(df_list){
  output <- list()
    for (i in 1:length(dfs_to_filter)){
      df_interest <- dfs_to_filter[[i]]
      if (nrow(df_interest) == 0) stop("one of your dataframes has no rows")
      not_i <- seq(1,length(dfs_to_filter))[seq(1,length(dfs_to_filter)) != i]
      for (j in not_i){
        df_filtered <- anti_join_rownames(df_interest, dfs_to_filter[[j]])
        if (nrow(df_filtered) == 0){
          break
        }
      }
      df_filtered <- subset(df_filtered, select = -c(Symbol)) 
      output <- c(output, list(df_filtered))
      print(dim(output)[[i]])
    }
  return(output)
}
```

#### Without Inhouse Data

```{r DE analysis without inhouse data}
register(MulticoreParam(4))
ddssva <- dds_no_inhouse
sv_names <- paste("SV", seq(1,svseq$n.sv), sep = "")
for (i in 1:length(sv_names)){
  colData(ddssva)[,sv_names[i]] <- svseq$sv[,i]
}
colData(ddssva)[,"Subpopulation"] <- as.factor(colData(ddssva)$Subpopulation)
colData(ddssva)[,"Tumor_JuxtaTumor"] <- as.factor(colData(ddssva)$Tumor_JuxtaTumor)
design(ddssva) <- ~ Subpopulation + Tumor_JuxtaTumor + SV1 + SV2 + SV3 + SV4 + SV5 + SV6 + SV7 + SV8 + SV9 + SV10 + SV11 + SV12 + SV13 + SV14 + SV15 + SV16 + SV17 + SV18 + SV19 + SV20
ddssva_copy <- ddssva
design(ddssva_copy) <- ~ Tumor_JuxtaTumor + SV1 + SV2 + SV3 + SV4 + SV5 + SV6 + SV7 + SV8 + SV9 + SV10 + SV11 + SV12 + SV13 + SV14 + SV15 + SV16 + SV17 + SV18 + SV19 + SV20 + SV21 + Subpopulation
#dds_no_inhouse_deseq <- DESeq(ddssva_copy)
#write_rds(dds_no_inhouse_deseq, file = "/home/kevin/Documents/PhD/subtypes/caf-subtype-analysis/dds_noinhouse_deseq_hgnc_25082022.Rds")
dds_no_inhouse_wald_diffdesign <- readRDS("dds_noinhouse_deseq_hgnc_25082022.Rds")
```

```{r All of these fail - model matrix less than full rank}
#design(dds_no_inhouse) <- ~ Subpopulation + Tumor_JuxtaTumor + SV1 + SV2 + SV3 + SV4
#design(dds_no_inhouse) <- ~ Tumor_JuxtaTumor + SV1 + SV2 + SV3 + SV4
#design(dds_no_inhouse) <- ~ SV1 + SV2 + SV3 + SV4
# design(dds_no_inhouse) <- ~ Tumor_JuxtaTumor + SV1 + SV2 + SV3 + SV4
```

The aim here is to extract lists of genes that are upregulated in each subpopulation compared to the other 2 subpopulations. To do this, steps from the [following tutorial](https://github.com/tavareshugo/tutorial_DESeq2_contrasts/blob/main/DESeq2_contrasts.md) were followed.

```{r Define model matrix 1}
# define model matrix
mod_mat <- model.matrix(design(dds_no_inhouse_wald_diffdesign), colData(dds_no_inhouse_wald_diffdesign))
# e.g. for each sample that is S1, get the mean of the coefficients all the components of the formula 
S1 <- colMeans(mod_mat[dds_no_inhouse_wald_diffdesign$Subpopulation == "S1",])
S3 <- colMeans(mod_mat[dds_no_inhouse_wald_diffdesign$Subpopulation == "S3",])
S4 <- colMeans(mod_mat[dds_no_inhouse_wald_diffdesign$Subpopulation == "S4",])
not_S1 <- colMeans(mod_mat[dds_no_inhouse_wald_diffdesign$Subpopulation %in% c("S3", "S4"),])
not_S3 <- colMeans(mod_mat[dds_no_inhouse_wald_diffdesign$Subpopulation %in% c("S1", "S4"),])
not_S4 <- colMeans(mod_mat[dds_no_inhouse_wald_diffdesign$Subpopulation %in% c("S1", "S3"),])
```

```{r Extract results for each pairwise contrast 1}
res_not_S1 <- results(dds_no_inhouse_wald_diffdesign, contrast = S1 - not_S1, filterFun = ihw, alpha = 0.05, lfcThreshold = 2, altHypothesis = "greater")
res_not_S3 <- results(dds_no_inhouse_wald_diffdesign, contrast = S3 - not_S3, filterFun = ihw, alpha = 0.05, lfcThreshold = 2, altHypothesis = "greater")
res_not_S4 <- results(dds_no_inhouse_wald_diffdesign, contrast = S4 - not_S4, filterFun = ihw, alpha = 0.05, lfcThreshold = 2, altHypothesis = "greater")
```

```{r}
drawLines <- function() abline(h=c(-2,2),col="dodgerblue",lwd=2)
par(mfrow = c(3, 1))
plotMA(res_not_S1) ; drawLines()
plotMA(res_not_S3) ; drawLines()
plotMA(res_not_S4) ; drawLines()
```

```{r Extract shrunken LFC for each contrast 1}
res_not_S1_shrink <- lfcShrink(dds_no_inhouse_wald_diffdesign, contrast = S1 - not_S1, res = res_not_S1, type = "ashr")
res_not_S3_shrink <- lfcShrink(dds_no_inhouse_wald_diffdesign, contrast = S3 - not_S3, res = res_not_S3, type = "ashr")
res_not_S4_shrink <- lfcShrink(dds_no_inhouse_wald_diffdesign, contrast = S4 - not_S4, res = res_not_S4, type = "ashr")
```

```{r MA plot shrunken LFC with 6144}
par(mfrow = c(3, 1))
plotMA(res_not_S1_shrink) ; drawLines()
plotMA(res_not_S3_shrink) ; drawLines()
plotMA(res_not_S4_shrink) ; drawLines()
```

```{r}
res_not_S1_shrink_extract <- get_upregulated(res_not_S1_shrink)
res_not_S3_shrink_extract <- get_upregulated(res_not_S3_shrink)
res_not_S4_shrink_extract <- get_upregulated(res_not_S4_shrink)
```

```{r}
dfs_to_filter <- list(res_not_S1_shrink_extract, res_not_S3_shrink_extract, res_not_S4_shrink_extract)
dfs_filtered <- filter_dfs_antijoin_rownames(dfs_to_filter)
names(dfs_filtered) <- c("S1", "S3", "S4")
s1_filtered <- dfs_filtered[[1]]
s1_filtered_annotations <- annotate_de_genes(s1_filtered, filter_by = "hgnc_symbol")
s3_filtered <- dfs_filtered[[2]]
s3_filtered_annotations <- annotate_de_genes(s3_filtered, filter_by = "hgnc_symbol")
s4_filtered <- dfs_filtered[[3]]
s4_filtered_annotations <- annotate_de_genes(s4_filtered, filter_by = "hgnc_symbol")
```

#### Without inhouse data or study EGAD00001006144

```{r DE analysis without inhouse or 6144}
sv_names <- paste("SV", seq(1,svseq_no6144$n.sv), sep = "")
for (i in 1:length(sv_names)){
  colData(dds_noinhouse_no6144)[,sv_names[i]] <- svseq_no6144$sv[,i]
}
colData(dds_noinhouse_no6144)[,"Subpopulation"] <- as.factor(colData(dds_noinhouse_no6144)$Subpopulation)
colData(dds_noinhouse_no6144)[,"Tumor_JuxtaTumor"] <- as.factor(colData(dds_noinhouse_no6144)$Tumor_JuxtaTumor)

design(dds_noinhouse_no6144) <- ~ Tumor_JuxtaTumor + SV1 + SV2 + SV3 + SV4 + SV5 + SV6 + SV7 + SV8 + SV9 + SV10 + SV11 + SV12 + SV13 + SV14 + SV15 + SV16 + SV17 + SV18 + Subpopulation
#ptm <- proc.time()
#dds_no_inhouse_no6144 <- DESeq(dds_noinhouse_no6144, parallel = TRUE)
#proc.time() - ptm
#write_rds(dds_no_inhouse_no6144, file = "/home/kevin/Documents/PhD/subtypes/caf-subtype-analysis/dds_noinhouse_no6144_hgnc_25082022.Rds")
dds_no_inhouse_no6144 <- readRDS("dds_noinhouse_no6144_hgnc_25082022.Rds")
```
Elapsed = 2835.825

```{r Run DESeq separately}
#register(MulticoreParam(4))
#ptm <- proc.time()
#dds_no_inhouse_no6144 <- DESeq(dds_noinhouse_no6144, parallel = TRUE)
#proc.time() - ptm
#write_rds(dds_no_inhouse_no6144, file = "/home/kevin/Documents/PhD/subtypes/caf-subtype-analysis/dds_no_inhouse_no6144_deseq_11082022.Rds")
#dds_no_inhouse_no6144 <- readRDS("dds_no_inhouse_no6144_deseq_11082022.Rds")
```

The aim here is to extract lists of genes that are upregulated in each subpopulation compared to the other 2 subpopulations. To do this, steps from the [following tutorial](https://github.com/tavareshugo/tutorial_DESeq2_contrasts/blob/main/DESeq2_contrasts.md) were followed.

```{r Define model matrix 2}
# define model matrix
mod_mat <- model.matrix(design(dds_no_inhouse_no6144), colData(dds_no_inhouse_no6144))
# e.g. for each sample that is S1, get the mean of the coefficients all the components of the formula 
S1 <- colMeans(mod_mat[dds_no_inhouse_no6144$Subpopulation == "S1",])
S3 <- colMeans(mod_mat[dds_no_inhouse_no6144$Subpopulation == "S3",])
S4 <- colMeans(mod_mat[dds_no_inhouse_no6144$Subpopulation == "S4",])
not_S1 <- colMeans(mod_mat[dds_no_inhouse_no6144$Subpopulation %in% c("S3", "S4"),])
not_S3 <- colMeans(mod_mat[dds_no_inhouse_no6144$Subpopulation %in% c("S1", "S4"),])
not_S4 <- colMeans(mod_mat[dds_no_inhouse_no6144$Subpopulation %in% c("S1", "S3"),])
```

```{r Extract results for each pairwise contrast 2}
res_not_S1 <- results(dds_no_inhouse_no6144, contrast = S1 - not_S1, filterFun = ihw, alpha = 0.05, lfcThreshold = 2, altHypothesis = "greater")
res_not_S3 <- results(dds_no_inhouse_no6144, contrast = S3 - not_S3, filterFun = ihw, alpha = 0.05, lfcThreshold = 2, altHypothesis = "greater")
res_not_S4 <- results(dds_no_inhouse_no6144, contrast = S4 - not_S4, filterFun = ihw, alpha = 0.05, lfcThreshold = 2, altHypothesis = "greater")
```

```{r}
drawLines <- function() abline(h=c(-2,2),col="dodgerblue",lwd=2)
par(mfrow = c(3, 1))
plotMA(res_not_S1) ; drawLines()
plotMA(res_not_S3) ; drawLines()
plotMA(res_not_S4) ; drawLines()
```

```{r Extract shrunken LFC for each contrast 2}
res_not_S1_shrink <- lfcShrink(dds_no_inhouse_wald_diffdesign, contrast = S1 - not_S1, res = res_not_S1, type = "ashr")
res_not_S3_shrink <- lfcShrink(dds_no_inhouse_wald_diffdesign, contrast = S3 - not_S3, res = res_not_S3, type = "ashr")
res_not_S4_shrink <- lfcShrink(dds_no_inhouse_wald_diffdesign, contrast = S4 - not_S4, res = res_not_S4, type = "ashr")
```

```{r MA plot shrunken LFC without 6144}
par(mfrow = c(3, 1))
plotMA(res_not_S1_shrink) ; drawLines()
plotMA(res_not_S3_shrink) ; drawLines()
plotMA(res_not_S4_shrink) ; drawLines()
```

```{r}
res_not_S1_shrink_extract <- get_upregulated(res_not_S1_shrink)
res_not_S3_shrink_extract <- get_upregulated(res_not_S3_shrink)
res_not_S4_shrink_extract <- get_upregulated(res_not_S4_shrink)
```

```{r}
dfs_to_filter <- list(res_not_S1_shrink_extract, res_not_S3_shrink_extract, res_not_S4_shrink_extract)
dfs_filtered <- filter_dfs_antijoin_rownames(dfs_to_filter)
names(dfs_filtered) <- c("S1", "S3", "S4")
s1_filtered <- dfs_filtered[[1]]
s1_filtered_annotations <- annotate_de_genes(s1_filtered, filter_by = "hgnc_symbol")
s3_filtered <- dfs_filtered[[2]]
s3_filtered_annotations <- annotate_de_genes(s3_filtered, filter_by = "hgnc_symbol")
s4_filtered <- dfs_filtered[[3]]
s4_filtered_annotations <- annotate_de_genes(s4_filtered, filter_by = "hgnc_symbol")
```

```{r Extract gene sets}
s1_gs <- s1_filtered_annotations$Gene
s3_gs <- s3_filtered_annotations$Gene
s4_gs <- s4_filtered_annotations$Gene
genesets <- list(s1_gs, s3_gs, s4_gs)
names(genesets) <- c("S1", "S3", "S4")
```

### Over-representation analysis

```{r Extract background Entrez IDs}
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", host = "https://uswest.ensembl.org")
                # host="www.ensembl.org")
background_genes <- getBM(attributes = "entrezgene_id", 
                          #filters = "ensembl_gene_id_version", 
                          filters = "hgnc_symbol",
                          mart = mart,
                          values = rownames(res_not_S1_shrink))
```


```{r ORA, echo=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)
ego_S1 <- enrichGO(gene       = s1_filtered_annotations$`Entrez ID`,
                universe      = as.character(background_genes$entrezgene_id),
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego_S3 <- enrichGO(gene          = s3_filtered_annotations$`Entrez ID`,
                universe      = as.character(background_genes$entrezgene_id),
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
ego_S4 <- enrichGO(gene       = s4_filtered_annotations$`Entrez ID`,
                universe      = as.character(background_genes$entrezgene_id),
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
goplot(ego_S1)

```

```{r}
dotplot(ego_S1, title = "S1 signature over-representation analysis")
```

```{r}
dotplot(ego_S3, title = "S3 signature over-representation analysis")
```

```{r}
dotplot(ego_S4, title = "S4 signature over-representation analysis")
```

### Gene set variation analysis (GSVA) for gene signature identification

```{r Get inhouse expression data}
idx <- which(dds$Study == "InHouse")
dds_inhouse <- dds[,idx]
```

```{r}
inhouse_metadata <- read.csv("/home/kevin/Documents/PhD/rna_seq_bc/metadata/reformat_samples_extra_info.csv")
caf_es <- gsva(expr = dds_inhouse,
               gset.idx.list = genesets,
               method = "gsva",
               kcdf = "Poisson",
               mx.diff=FALSE)
caf_es$Patient <- inhouse_metadata$Patient
caf_es$Subtype <- inhouse_metadata$Subtype
caf_es$Grade <- inhouse_metadata$Grade
caf_es$Histology <- inhouse_metadata$Histology
```

```{r Plot heatmaps}
subpopulationOrder <- c("tumor", "juxtatumor")
sampleOrderBySubpopulation <- sort(match(caf_es$Tumor_JuxtaTumor, subpopulationOrder),
                             index.return=TRUE)$ix
subpopulationXtable <- table(caf_es$Tumor_JuxtaTumor)
subpopulationColorLegend <- c(tumor="red", juxtatumor="green")
geneSetOrder <- c("S4", "S3", "S1")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(caf_es)[geneSetOrder, sampleOrderBySubpopulation], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(subpopulationColorLegend[subpopulationOrder],
                          times=subpopulationXtable[subpopulationOrder]),
        labCol="", 
        caf_es$Tumor_JuxtaTumor[sampleOrderBySubpopulation],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.285,1.1, "CAF", col="red", cex=1.2)
text(0.55,1.1, "TAN", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)

subtypeOrder <- c("LuminalA", "TNBC")
sampleOrderBySubtype <- sort(match(caf_es$Subtype, subtypeOrder),
                             index.return=TRUE)$ix
subtypeXtable <- table(caf_es$Subtype)
subtypeColorLegend <- c(LuminalA="red", TNBC="green")
geneSetOrder <- c("S4", "S3", "S1")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(caf_es)[geneSetOrder, sampleOrderBySubtype], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(subtypeColorLegend[subtypeOrder],
                          times=subtypeXtable[subtypeOrder]),
        labCol="", 
        #labCol = caf_es$Patient[sampleOrderBySubtype],
        caf_es$Subtype[sampleOrderBySubtype],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.4,1.1, "LuminalA", col="red", cex=1.2)
text(0.65,1.1, "TNBC", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)

gradeOrder <- c("Grade_2", "Grade_3")
sampleOrderByGrade <- sort(match(caf_es$Grade, gradeOrder),
                             index.return=TRUE)$ix
gradeXtable <- table(caf_es$Grade)
gradeColorLegend <- c(Grade_2="red", Grade_3="green")
geneSetOrder <- c("S4", "S3", "S1")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(caf_es)[geneSetOrder, sampleOrderByGrade], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(gradeColorLegend[gradeOrder],
                          times=gradeXtable[gradeOrder]),
        labCol="", 
        #labCol = caf_es$Patient[sampleOrderByGrade],
        caf_es$Subtype[sampleOrderBySubtype],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.34,1.1, "Grade 2", col="red", cex=1.2)
text(0.6,1.1, "Grade 3", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)
```

```{r}
subtypeOrder <- c("LuminalA", "TNBC")
sampleOrderBySubtype <- sort(match(caf_es$Subtype, subtypeOrder),
                             index.return=TRUE)$ix
subtypeXtable <- table(caf_es$Subtype)
subtypeColorLegend <- c(LuminalA="red", TNBC="green")
geneSetOrder <- c("S4", "S3", "S1")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(caf_es)[geneSetOrder, sampleOrderBySubtype], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(subtypeColorLegend[subtypeOrder],
                          times=subtypeXtable[subtypeOrder]),
        #labCol="", 
        labCol = caf_es$Patient[sampleOrderBySubtype],
        caf_es$Subtype[sampleOrderBySubtype],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.4,1.1, "LuminalA", col="red", cex=1.2)
text(0.65,1.1, "TNBC", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)

```

```{r}
caf_idx <- which(colData(caf_es)$Tumor_JuxtaTumor == "tumor")
tan_idx <- which(colData(caf_es)$Tumor_JuxtaTumor == "juxtatumor")
df_plot_S1 <- cbind.data.frame(S1_ES = assay(caf_es)[1,], Tumor_JuxtaTumor = colData(caf_es)$Tumor_JuxtaTumor)
gsva_plot_s1 <- ggplot(df_plot_S1, aes(x = Tumor_JuxtaTumor, y = S1_ES)) +
  geom_point(size = 2,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.1,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ylab("S1 GSVA ES")  #, axis.title.x = element_blank())

df_plot_S3 <- cbind.data.frame(S3_ES = assay(caf_es)[2,], Tumor_JuxtaTumor = colData(caf_es)$Tumor_JuxtaTumor)
gsva_plot_s3 <- ggplot(df_plot_S3, aes(x = Tumor_JuxtaTumor, y = S3_ES)) +
  geom_point(size = 2,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.1,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ylab("S3 GSVA ES")  #, axis.title.x = element_blank())

df_plot_S4 <- cbind.data.frame(S4_ES = assay(caf_es)[3,], Tumor_JuxtaTumor = colData(caf_es)$Tumor_JuxtaTumor)
gsva_plot_s4 <- ggplot(df_plot_S4, aes(x = Tumor_JuxtaTumor, y = S4_ES)) +
  geom_point(size = 2,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.1,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ylab("S4 GSVA ES")  #, axis.title.x = element_blank())
```

```{r}
distribution_es_s1 <- df_plot_S1 %>% 
ggplot(mapping = aes(S1_ES)) +
  geom_density() +
  facet_wrap(~Tumor_JuxtaTumor) +
  theme_linedraw()

distribution_es_s3 <- df_plot_S3 %>% 
ggplot(mapping = aes(S3_ES)) +
  geom_density() +
  facet_wrap(~Tumor_JuxtaTumor) +
  theme_linedraw()

distribution_es_s4 <- df_plot_S4 %>% 
ggplot(mapping = aes(S4_ES)) +
  geom_density() +
  facet_wrap(~Tumor_JuxtaTumor) +
  theme_linedraw()
```

```{r}
top_row <- plot_grid(gsva_plot_s1, gsva_plot_s3, gsva_plot_s4,
      ncol = 3,
      labels = c('A', 'B', 'C'),
      label_fontfamily = 'serif',
      label_fontface = 'bold',
      label_size = 15,
      align = 'h',
      rel_widths = c(1.10, 0.80, 1.10))

    bottom_row <- plot_grid(distribution_es_s1, distribution_es_s3, distribution_es_s4,
      ncol = 3,
      labels = c('D', 'E', 'F'),
      label_fontfamily = 'serif',
      label_fontface = 'bold',
      label_size = 22,
      align = 'h',
      rel_widths = c(0.8, 1.2))

    plot_grid(top_row, bottom_row, nrow = 2,
      rel_heights = c(1.1, 0.9))
```

```{r Shapiro Wilk test for normality}
df_plot_S1$S1_ES %>% 
shapiro.test()

df_plot_S3$S3_ES %>% 
shapiro.test()

df_plot_S4$S4_ES %>% 
shapiro.test()
```

p < 0.05, data is not normally distributed. T-test may be unsuitable for comparing distributions. Carry out Wilcoxon rank-sum test (Mann-Whitney U test).

```{r}
wilcox.test(x = df_plot_S1$S1_ES[df_plot_S1$Tumor_JuxtaTumor == "tumor"], 
            y = df_plot_S1$S1_ES[df_plot_S1$Tumor_JuxtaTumor == "juxtatumor"],
            paired = TRUE)
wilcox.test(x = df_plot_S3$S3_ES[df_plot_S3$Tumor_JuxtaTumor == "tumor"], 
            y = df_plot_S3$S3_ES[df_plot_S3$Tumor_JuxtaTumor == "juxtatumor"],
            paired = TRUE)
wilcox.test(x = df_plot_S4$S4_ES[df_plot_S4$Tumor_JuxtaTumor == "tumor"], 
            y = df_plot_S4$S4_ES[df_plot_S4$Tumor_JuxtaTumor == "juxtatumor"],
            paired = TRUE)
```

```{r Find medians of enrichment scores in S1 and S3 in CAF and TAN}
median(df_plot_S1$S1_ES[df_plot_S1$Tumor_JuxtaTumor == "tumor"])
median(df_plot_S1$S1_ES[df_plot_S1$Tumor_JuxtaTumor == "juxtatumor"])
median(df_plot_S3$S3_ES[df_plot_S3$Tumor_JuxtaTumor == "tumor"])
median(df_plot_S3$S3_ES[df_plot_S3$Tumor_JuxtaTumor == "juxtatumor"])
```

We can see from the above that there is a difference between the enrichment scores for the gene signatures for the S1 and S3 subpopulaions between CAF and TAN. We can suggest that the TAN (juxta-tumour) samples are enriched for our S1 and S3 gene signatures compared to the genes outside the gene signatures, and that this enrichment is greater in the TAN samples than in the CAF samples.

```{r}
results_combined <- cbind.data.frame(colData(caf_es)$names, colData(caf_es)$Patient, df_plot_S1$S1_ES, df_plot_S3$S3_ES, df_plot_S4$S4_ES, df_plot_S1$Tumor_JuxtaTumor, colData(caf_es)$Subtype, colData(caf_es)$Grade, colData(caf_es)$Histology)
colnames(results_combined) <- c("Sample", "Patient", "S1_ES", "S3_ES", "S4_ES", "Tumor_JuxtaTumor", "Subtype", "Grade", "Histology")
results_combined
```

### Chemoresistance gene signature

Su et al 2018 carried out differential expression analysis between chemoresistant and chemosensitive CAFs from breast cancer. It is possible to try and identify the resulting signature in our samples. We will use the list of upregulated genes in chemoresistant samples as our marker of chemoresistance and the genes downregulated in the chemoresistant cells as our marker of chemosensitivity.

The HGNCHelper package can be used to fix outdated gene symbols.

```{r read in gene signatures}
gene_signature <- read.csv("/home/kevin/Documents/PhD/rna_seq_bc/gene_signature/1-s2.0-S0092867418300448-mmc1.csv", skip = 1)
gene_signature_loc <- gene_signature[str_detect(gene_signature$Gene.name, pattern = "^LOC\\d+$"),]
# write to file and look up loc gene signatures manually, read back in as gene_signature_loc_official
#write.table(gene_signature_loc, file = "/home/kevin/Documents/PhD/rna_seq_bc/gene_signature/chemo_signature_loc.txt", quote = F, row.names = F)
gene_signature_loc_official <- read.table("/home/kevin/Documents/PhD/rna_seq_bc/gene_signature/chemo_signature_loc_official_names.txt", header = T)
gene_signature_proper_names <- gene_signature
gene_signature_proper_names$Official_name <- gene_signature$Gene.name
idx <- match(gene_signature_loc$Gene.name, gene_signature$Gene.name)
gene_signature_proper_names[idx,]$Official_name <- gene_signature_loc_official$Official_symbol
gene_signature_proper_names <- drop_na(gene_signature_proper_names)
row_remove <- which(gene_signature_proper_names$Gene.name == "N/A")
gene_signature_proper_names <- gene_signature_proper_names[-c(row_remove),]
gene_signature_hgnc <- checkGeneSymbols(gene_signature_proper_names$Official_name, species = "human")
gene_signature_hgnc$Suggested.Symbol[which(gene_signature_hgnc$x == "CCRL1")] <- NA
gene_signature_hgnc$Suggested.Symbol[which(gene_signature_hgnc$x == "FLJ40504")] <- NA
gene_signature_hgnc$Suggested.Symbol[which(gene_signature_hgnc$x == "LETR1")] <- "LETR1"
gene_signature_hgnc <- drop_na(gene_signature_hgnc)
colnames(gene_signature_hgnc) <- c("Gene.name", "Approved", "Suggested.Symbol")
gene_signature_join <- full_join(gene_signature, gene_signature_hgnc, by = "Gene.name")
gene_signature_chemoresistance <- gene_signature_join$Suggested.Symbol[which(gene_signature_join$Regulation == "up")]
gene_signature_chemoresistance <- gene_signature_chemoresistance[!is.na(gene_signature_chemoresistance)]
gene_signature_chemosensitivity <- gene_signature_join$Suggested.Symbol[which(gene_signature_join$Regulation == "down")]
gene_signature_chemosensitivity<- gene_signature_chemosensitivity[!is.na(gene_signature_chemosensitivity)]
genesets_chemo <- list(gene_signature_chemoresistance, gene_signature_chemosensitivity)
names(genesets_chemo) <- c("Chemoresistance", "Chemosensitive")
```

```{r GSVA for chemoresistance signatures}
chemo_es <- gsva(expr = dds_inhouse,
               gset.idx.list = genesets_chemo,
               method = "gsva",
               kcdf = "Poisson",
               mx.diff=FALSE)
chemo_es$Patient <- inhouse_metadata$Patient
chemo_es$Subtype <- inhouse_metadata$Subtype
chemo_es$Grade <- inhouse_metadata$Grade
chemo_es$Histology <- inhouse_metadata$Histology
```
```{r}
assay(chemo_es)
```

```{r}
subpopulationOrder <- c("tumor", "juxtatumor")
sampleOrderBySubpopulation <- sort(match(chemo_es$Tumor_JuxtaTumor, subpopulationOrder),
                             index.return=TRUE)$ix
subpopulationXtable <- table(chemo_es$Tumor_JuxtaTumor)
subpopulationColorLegend <- c(tumor="red", juxtatumor="green")
geneSetOrder <- c("Chemoresistance", "Chemosensitive")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(chemo_es)[geneSetOrder, sampleOrderBySubpopulation], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(subpopulationColorLegend[subpopulationOrder],
                          times=subpopulationXtable[subpopulationOrder]),
        labCol="", 
        chemo_es$Tumor_JuxtaTumor[sampleOrderBySubpopulation],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.285,1.1, "CAF", col="red", cex=1.2)
text(0.55,1.1, "TAN", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)

subtypeOrder <- c("LuminalA", "TNBC")
sampleOrderBySubtype <- sort(match(chemo_es$Subtype, subtypeOrder),
                             index.return=TRUE)$ix
subtypeXtable <- table(chemo_es$Subtype)
subtypeColorLegend <- c(LuminalA="red", TNBC="green")
geneSetOrder <- c("Chemoresistance", "Chemosensitive")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(chemo_es)[geneSetOrder, sampleOrderBySubtype], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(subtypeColorLegend[subtypeOrder],
                          times=subtypeXtable[subtypeOrder]),
        labCol="", 
        #labCol = caf_es$Patient[sampleOrderBySubtype],
        chemo_es$Subtype[sampleOrderBySubtype],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.4,1.1, "LuminalA", col="red", cex=1.2)
text(0.65,1.1, "TNBC", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)

gradeOrder <- c("Grade_2", "Grade_3")
sampleOrderByGrade <- sort(match(chemo_es$Grade, gradeOrder),
                             index.return=TRUE)$ix
gradeXtable <- table(chemo_es$Grade)
gradeColorLegend <- c(Grade_2="red", Grade_3="green")
geneSetOrder <- c("Chemoresistance", "Chemosensitive")
geneSetLabels <- geneSetOrder
hmcol <- colorRampPalette(brewer.pal(10, "RdBu"))(256)
hmcol <- hmcol[length(hmcol):1]
heatmap(assay(chemo_es)[geneSetOrder, sampleOrderByGrade], Rowv=NA,
        Colv=NA, scale="row", margins=c(3,5), col=hmcol,
        ColSideColors=rep(gradeColorLegend[gradeOrder],
                          times=gradeXtable[gradeOrder]),
        labCol="", 
        #labCol = caf_es$Patient[sampleOrderByGrade],
        chemo_es$Subtype[sampleOrderBySubtype],
        labRow=paste(toupper(substring(geneSetLabels, 1,1)),
                     substring(geneSetLabels, 2), sep=""),
        cexRow=2, main=" \n ")
par(xpd=TRUE)
text(0.34,1.1, "Grade 2", col="red", cex=1.2)
text(0.6,1.1, "Grade 3", col="green", cex=1.2)
#text(0.47,1.21, "S4", col="blue", cex=1.2)
mtext("Gene sets", side=4, line=0, cex=1.5)
mtext("Samples", side=1, line=4, cex=1.5, at = 0.42)
```

```{r}
chemoresistance_ratio <- function(df){
  out <- data.frame(Patient = df$Patient,)
}
as_tibble(t(assay(chemo_es))) %>% mutate(Patient = colData(chemo_es)$Patient, Chemoresistance_ratio = abs(Chemoresistance)/abs(Chemosensitive))
```


### Marker gene expression

Source: GeneCards

 -  ITGB1 (CD29) is an integrin, integrins are involved in cell-cell and cell-ECM adhesion. It is expressed in 3 of the 4 subpopulations (not S2).
 -  FAP is Fibroblast activation protein, a serine proteinase, selectively expressed in reactive stromal fibroblasts in epithelial cancers, expression should be highest in S1 population. Roles include: tissue remodelling, fibrosis, wound healing.
 -  α-SMA - ACTA2, alpha smooth muscle actin is involved in cell motility, structure, integrity and signalling. Its expression is highest in S1 and S4 CAFs.
 -  PDPN (podoplanin) increases cell motility. It should only be expressed in S1 cells.
 -  PDGFRβ (Platelet Derived Growth Factor Receptor Beta) is a tyrosine kinase receptor for platelet-derived growth factors. They stimulate mitosis of cells of mesenchymal origin. Expression should be highest in S1, some expression in S3 and S4.
 
There are other possible genes which could be used to distinguish them through manual curation: FSP1, CAV1, DPP4

```{r}
# want to create 'background' gene set entrez id + LFC values for all genes
#info <- getBM(attributes = c("hgnc_symbol", "ensembl_gene_id_version"),
 #             mart=mart, filters = "ensembl_gene_id_version", values = rownames(dds))
info2 <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
              mart=mart, filters = "hgnc_symbol", values = str_split_fixed(string = rownames(dds), pattern ="\\.", n = 2)[,1])
```

```{r Plotting functions}
create_df_plotcounts <- function(dds, gene_interest, info, symbol_type_dds = "hgnc_symbol"){
  library(DESeq2)
  library(biomaRt)
  # put check for type of gene symbol in dds object, assuming ensembl gene version atm
  # assume info have column called hgnc_symbol
  info = info[!duplicated(info$hgnc_symbol),]
  rownames(info) = info$hgnc_symbol
  # assuming genes interest are in info, put in check for this to deal with missing gene symbol
  if (!(gene_interest %in% info$hgnc_symbol)){
    print(paste(gene_interest, "not found in table of hgnc symbols", sep = " "))
    geneCounts <- NA
  } else {
    gene_interest_info <- info[gene_interest,]
  #print(gene_interest_info)
    if (symbol_type_dds == "ensembl_gene_id_version"){
      gene_ensg <- info$ensembl_gene_id[which(info$hgnc_symbol == gene_interest)]
  # find ENSG1234.1 in dds object given ENGS1234
      idx <- str_detect(rownames(assay(dds)), paste0(gene_ensg, "\\.."))
      gene_correct_version <- rownames(assay(dds))[idx]
  } else if (symbol_type_dds == "hgnc_symbol"){
    #idx <- str_detect(rownames(assay(dds)), gene_interest)
    #gene_correct_version <- rownames(assay(dds))[idx]
      gene_correct_version <- gene_interest
  } else {
    stop("symbol_type_dds must be one of hgnc_symbol or ensembl_gene_id_version")
  }
    geneCounts <- plotCounts(dds, gene = gene_correct_version, intgroup = c("Study", "Subpopulation", "Tumor_JuxtaTumor"), returnData = TRUE)
  }
}


caf_plot_tumour_juxtatumour <- function(df_for_plotting, gene = NULL){
  library(ggplot2)
  gene = gene
  if (is.list(df_for_plotting)){
     cols <- colnames(df_for_plotting)
     df_for_plotting <- as.data.frame(df_for_plotting)
     colnames(df_for_plotting) <- cols
  }
  ggplot(df_for_plotting, aes(x = Subpopulation, y = count,  colour = Tumor_JuxtaTumor)) +
  geom_point(size = 1,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  scale_y_log10() +
  labs(title = gene) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ylab("Normalised counts")  #, axis.title.x = element_blank())
}

caf_plot_study <- function(df_for_plotting, gene = NULL){
  library(ggplot2)
  gene = gene
  if (is.list(df_for_plotting)){
     cols <- colnames(df_for_plotting)
     df_for_plotting <- as.data.frame(df_for_plotting)
     colnames(df_for_plotting) <- cols
  }
  ggplot(df_for_plotting, aes(x = Subpopulation, y = count,  colour = Study)) +
  geom_point(size = 1,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  ) +
  scale_y_log10() +
  labs(title = gene) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"), 
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ylab("Normalised counts")  #, axis.title.x = element_blank())
}

```

```{r Plot CAF marker genes}
genes_interest <- c("FAP", "ITGB1", "ACTA2", "PDPN", "PDGFRB")
genes_interest_common_names <- c("FAP", "CD29", "αSMA", "PDPN", "PDGFRB")
annotated_dfs_for_plotting <- lapply(X = genes_interest, FUN = create_df_plotcounts, dds = dds, info = info2)
names(annotated_dfs_for_plotting) <- genes_interest
plots_out_tumor_juxtatumor <- list()
for (i in 1:length(annotated_dfs_for_plotting)){
  plt <- caf_plot_tumour_juxtatumour(df_for_plotting = annotated_dfs_for_plotting[[i]], 
                                                                   gene = genes_interest_common_names[i])
  plots_out_tumor_juxtatumor[[i]] = plt
}
plots_out_study <- list()
for (i in 1:length(annotated_dfs_for_plotting)){
  plt <- caf_plot_study(df_for_plotting = annotated_dfs_for_plotting[[i]], 
                        gene = genes_interest_common_names[i])
  plots_out_study[[i]] = plt
}
ggar_obj_tumor_juxtatumor <- ggarrange(plotlist = plots_out_tumor_juxtatumor, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_tumor_juxtatumor_annotated <- annotate_figure(ggar_obj_tumor_juxtatumor, bottom = text_grob("CAF Subpopulation"))
ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"))
ggar_obj_tumor_juxtatumor_annotated
ggar_obj_study_annotated
```

```{r Plot some other genes of interest}
genes_interest <- c("CXCL12","TNFSF4","PDCD1LG2", "CD276", "NT5E", "DPP4", "CAV1", "ATL1")
genes_interest_common_names <- c("CXCL12", "OX40L", "PDL2", "B7H3", "CD73", "DPP4", "CAV1", "FSP1")
annotated_dfs_for_plotting <- lapply(X = genes_interest, FUN = create_df_plotcounts, dds = dds, info = info2)
names(annotated_dfs_for_plotting) <- genes_interest
plots_out_tumor_juxtatumor <- list()
for (i in 1:length(annotated_dfs_for_plotting)){
  if (is.null(dim(annotated_dfs_for_plotting[[i]]))){
    next
  } else {
    plt <- caf_plot_tumour_juxtatumour(df_for_plotting = annotated_dfs_for_plotting[[i]], 
                                                    gene = genes_interest_common_names[i])
    plots_out_tumor_juxtatumor[[i]] = plt
  }
}
plots_out_tumor_juxtatumor <- plots_out_tumor_juxtatumor[lengths(plots_out_tumor_juxtatumor) != 0]
plots_out_study <- list()
for (i in 1:length(annotated_dfs_for_plotting)){
  if (is.null(dim(annotated_dfs_for_plotting[[i]]))){
    next
  } else {
    plt <- caf_plot_study(df_for_plotting = annotated_dfs_for_plotting[[i]], 
                        gene = genes_interest_common_names[i])
    plots_out_study[[i]] = plt
  }
}
plots_out_study <- plots_out_study[lengths(plots_out_study) != 0]
ggar_obj_tumor_juxtatumor <- ggarrange(plotlist = plots_out_tumor_juxtatumor, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_tumor_juxtatumor_annotated <- annotate_figure(ggar_obj_tumor_juxtatumor, bottom = text_grob("CAF Subpopulation"))
ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"))
ggar_obj_tumor_juxtatumor_annotated
ggar_obj_study_annotated
```

```{r Plot haemopoietic epithelial and endothelial markers - should be low}
# plotting original counts here, transformation 
genes_interest <- c("PTPRC", "EPCAM", "PECAM1")
genes_interest_common_names <- c("CD45", "EPCAM", "CD31")
annotated_dfs_for_plotting <- lapply(X = genes_interest, FUN = create_df_plotcounts, dds = dds, info = info2)
names(annotated_dfs_for_plotting) <- genes_interest
plots_out_tumor_juxtatumor <- list()
for (i in 1:length(annotated_dfs_for_plotting)){
  plt <- caf_plot_tumour_juxtatumour(df_for_plotting = annotated_dfs_for_plotting[[i]], 
                                                                   gene = genes_interest_common_names[i])
  plots_out_tumor_juxtatumor[[i]] = plt
}
plots_out_study <- list()
for (i in 1:length(annotated_dfs_for_plotting)){
  plt <- caf_plot_study(df_for_plotting = annotated_dfs_for_plotting[[i]], 
                        gene = genes_interest_common_names[i])
  plots_out_study[[i]] = plt
}
ggar_obj_tumor_juxtatumor <- ggarrange(plotlist = plots_out_tumor_juxtatumor, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_tumor_juxtatumor_annotated <- annotate_figure(ggar_obj_tumor_juxtatumor, bottom = text_grob("CAF Subpopulation"))
ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"))
print(ggar_obj_tumor_juxtatumor_annotated)
print(ggar_obj_study_annotated)
```

The low to zero counts of these genes are what we would expect.

Per Gene, do the data line up with what we would expect?
 -  FAP, S1 = high, would expect it to be lower in S3 than S4, they are roughly the same. Unknown Subpopulation, high, tight confidence
 -  ITGB1 (CD29), very similar across all subpopulations This is as expected - it is thought to not be expressed only in S2 which we don't have.
 -  ACTA2 (alphaSMA), very similar across all subpopulations. We would expect the expression to be lowest in the S3 subpopulation but we don't see that here.
 -  PDPN - expression highest in unknown samples, similar but lower in other subpopulations. Slightly higher in S1 than S3 and S4 - supposed to be expressed only in S1! Post-transcriptional regulation??? Seems to undergo post-translational regulation - https://www.spandidos-publications.com/10.3892/ijo.2013.1887#b53-ijo-42-06-1849 it is possible that differences won't be seen here on the mRNA level.
 -  PDGFRB - highest in S1, as expected. Expression in our unknown similar to S3 and S4.


 -  Carry out batch correction on reduced dataset, tumour_juxtatumour and subpopulation as model matrix
 -  DE analysis

### Batch correction with Combat-Seq

This batch correction allows us to remove differences between studies in the Mechta-Grigoriou CAF subpopulation data (i.e. excluding our own data), while still preserving differences between CAFs and TANs (tumour and juxta-tumour). This will allow us to carry out DE analysis on these samples.

```{r}
#counts_matrix <- assay(dds_no_inhouse_copy)
#batch <- metadata_no_inhouse$Study
#covariates <- metadata_no_inhouse$Tumor_JuxtaTumor
#adjusted <- ComBat_seq(counts = counts_matrix, batch = batch, group = covariates)
#dds_batch_corrected_noinhouse <- DESeqDataSetFromMatrix(adjusted, colData = metadata_no_inhouse, design = ~1)
```

```{r}
#ensembl_ids <- rownames(adjusted)
#ensembl_ids <- str_split_fixed(ensembl_ids, pattern = "\\.", n = 2)[,1]
#info <- getBM(attributes=c("hgnc_symbol",
#                           "ensembl_gene_id"),
#                  filters = c("ensembl_gene_id"),
#                  values = ensembl_ids,
#                  mart = mart,
 #                 useCache=FALSE)
```

```{r}
#genes_interest <- c("FAP", "ITGB1", "ACTA2", "PDPN", "PDGFRB")
#genes_interest_common_names <- c("FAP", "CD29", "αSMA", "PDPN", "PDGFRB")

#annotated_dfs_for_plotting <- lapply(X = genes_interest, FUN = create_df_plotcounts, dds = dds_batch_corrected_noinhouse)
#names(annotated_dfs_for_plotting) <- genes_interest
#plots_out_tumor_juxtatumor <- list()
#for (i in 1:length(annotated_dfs_for_plotting)){
#  plt <- caf_plot_tumour_juxtatumour(df_for_plotting = annotated_dfs_for_plotting[[i]], 
#                                                                   gene = genes_interest_common_names[i])
#  plots_out_tumor_juxtatumor[[i]] = plt
#}
#plots_out_study <- list()
#for (i in 1:length(annotated_dfs_for_plotting)){
#  plt <- caf_plot_study(df_for_plotting = annotated_dfs_for_plotting[[i]], 
#                        gene = genes_interest_common_names[i])
#  plots_out_study[[i]] = plt
#}
#ggar_obj_tumor_juxtatumor <- ggarrange(plotlist = plots_out_tumor_juxtatumor, common.legend = TRUE) # rel_heights values control title margins
#title <- ggdraw() + draw_label("Batch-corrected normalised expression of marker genes\naccording to CAF subpopulation", fontface='bold')
#ggar_obj_tumor_juxtatumor_annotated <- annotate_figure(ggar_obj_tumor_juxtatumor, bottom = text_grob("CAF Subpopulation"), top = text_grob("Batch-corrected normalised expression of marker genes\naccording to CAF subpopulation", color = "black", face = "bold", size = 10))
#ggar_obj_tumor_juxtatumor_annotated <- annotate_figure(ggar_obj_tumor_juxtatumor, bottom = text_grob("CAF Subpopulation"))
#ggar_obj_study <- ggarrange(plotlist = plots_out_study, common.legend = TRUE) # rel_heights values control title margins
#ggar_obj_study_annotated <- annotate_figure(ggar_obj_study, bottom = text_grob("CAF Subpopulation"), top = text_grob("Batch-corrected normalised expression of marker genes\naccording to CAF subpopulation", color = "black", face = "bold", size = 10))
#print(ggar_obj_tumor_juxtatumor_annotated)
#print(ggar_obj_study_annotated)
```

 -  Carry out batch correction between studies, cannot account for differences in subpopulations here
 -  Differential expression analysis CAF vs TAN
 -  We cannot include patient in the formula here as we don't have matched CAF TAN
 




```{r}
# have to divide everything by 10 because the max digit is ~10x the max machine integer
dds_adjusted_all <- DESeqDataSetFromMatrix(countData = round(adjusted_all/10), colData = metadata, design = ~ Study + Tumor_JuxtaTumor)
dds_adjusted_all$Tumor_JuxtaTumor <- relevel(dds_adjusted_all$Tumor_JuxtaTumor, ref = "juxtatumor")
dds_adjusted_all <- DESeq(dds_adjusted_all)
```

### Differentially Expressed Genes CAF vs TAN (Tumor vs Juxta-Tumor)

```{R, message=F, warning=F}
dds <- dds_adjusted_all
res <- results(dds, filterFun = ihw, alpha = 0.05, c("Tumor_JuxtaTumor", "tumor", "juxtatumor"))
lfc <- lfcShrink(dds = dds, res = res, coef = 6, type = "apeglm")
lfc_df <- as.data.frame(lfc)

up <- get_upregulated(lfc_df)
down <- get_downregulated(lfc_df)

up <- annotate_de_genes(up, filter_by = "ensembl_gene_id_version")
down <- annotate_de_genes(down, filter_by = "ensembl_gene_id_version")
```

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
master_lfc1 <- rbind(up,down)
# want to create 'background' gene set entrez id + LFC values for all genes
info <- getBM(attributes = c("hgnc_symbol", "entrezgene_id", "ensembl_gene_id_version"),
              mart=mart, filters = "hgnc_symbol", values = rownames(lfc_df))

lfc_df$Gene <- rownames(lfc_df)
tmp <- merge(info, lfc_df, by.x="hgnc_symbol", by.y="Gene")

background <- tmp$log2FoldChange
names(background) <- tmp$hgnc_symbol
background <- sort(background, decreasing = TRUE)

# make sure you have all DEGs here (LFC > 1 and pval cutoff i.e maser_lfc1 genes)
degtmp <- subset(tmp, tmp$hgnc_symbol %in% master_lfc1$Gene)
deg <- degtmp$entrezgene_id
deg <- na.omit(deg)

#deg.df <- bitr(deg, fromType = "ENTREZID",
 #               toType = c("ENSEMBL", "SYMBOL"),
  #              OrgDb = org.Hs.eg.db)
```

#### Hallmarks GSEA

```{r}
hmarks <- read.gmt("~/Downloads/h.all.v7.4.symbols.gmt")

egmt <- GSEA(background,
             TERM2GENE=hmarks,
             pvalueCutoff = 0.1, 
             pAdjustMethod = 'BH',
             minGSSize = 1, 
             maxGSSize = 1000,
             by='fgsea',
             nPermSimple = 10000,
             seed=11384191)

egmt_df <- egmt@result

egmt_subs <- subset(egmt_df, select=c(Description, enrichmentScore, NES, pvalue, p.adjust, core_enrichment ))

#DT::datatable(egmt_subs, rownames = FALSE, options=list(scrollX=T))
```

```{R, message=F, warning=F, fig.height=5, fig.width=7}
dotplot(egmt, title = "GSEA HALLMARKS", x="NES")
```

### Enrichment plots

```{R, message=F, warning=F, fig.width=10, fig.height=5}
library(enrichplot)
for(i in 1:5){
  p <- gseaplot2(egmt, geneSetID = i, title = egmt_df$Description[[i]], pvalue_table = T)
  plot(p)
}
```

#### GO Biological Processes GSEA

```{R,message=F, warning=F}
gobp <- read.gmt("~/Downloads/c5.go.bp.v7.4.symbols.gmt")

egmt <- GSEA(background,
             TERM2GENE=gobp,
             pvalueCutoff = 0.1, 
             pAdjustMethod = 'BH',
             minGSSize = 1, 
             maxGSSize = 1000,
             by='fgsea',
             nPermSimple = 10000,
             seed=11384191)

egmt_df <- egmt@result
#egmt_df$Description <- gsub("GOBP_", "", egmt_df$Description)
#egmt_df$ID <- gsub("GOBP_", "", egmt_df$ID)

egmt_subs <- subset(egmt_df, select=c(Description, enrichmentScore, NES, pvalue, p.adjust, core_enrichment ))
egmt_subs[1:5,]
#DT::datatable(egmt_subs, rownames = FALSE, options=list(scrollX=T))
```

### Dot Plot

```{R, message=F, warning=F, fig.height=10, fig.width=12}
dotplot(egmt, title = "GO BIOLOGICAL PROCESSES", x="NES", showCategory=30, font.size = 8)
```

### Enrichment Plots

```{R, message=F, warning=F, fig.width=10, fig.height=5}
for(i in 1:5){
  p <- gseaplot2(egmt, geneSetID = i, title = egmt_df$Description[[i]], pvalue_table = T)
  plot(p)
}
```


### Batch correction

The following graphs demonstrate the batch effects

```{r}
ann_colors = list(Study=c(EGAD00001004810 = "forestgreen",  EGAD00001003808 = "gold", EGAD00001006144 = "blue", EGAD00001005744 = "magenta3", InHouse = "black"),
                  Subpopulation = c(S1 = "dodgerblue4", S3 = "chartreuse1", S4 = "grey67", Unknown = "palevioletred"),
                  Tumor_JuxtaTumor = c(tumor = "royalblue", juxtatumor = "red")
                  )
```

```{r Generate distance matrix + generate heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
sampleDists <- dist(t(assay(vsd)))
# don't want files column to be on heatmap
subset_coldata <- subset(coldata, select = -c(files))
sampleDistMatrix <- as.matrix( sampleDists )
rownames(subset_coldata) <- subset_coldata$names
subset_coldata$names <- NULL

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```


```{r}
pcaData <- plotPCA(vsd, intgroup = c("Study", "Subpopulation"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData, aes(x = PC1, y = PC2, color = Study, shape = Subpopulation)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with VST data")
```

Next step is to look at batch correction.

#### Method 1: Limma's remove batch effect function

```{r message=FALSE}
library(limma)
```

```{r create copies of vsd to avoid confusion}
vsd_remove_batch_intercept <- vsd
vsd_remove_batch_tumor_juxta_subpopulation <- vsd
vsd_remove_batch_tumor_juxta_only <- vsd
```


```{r removeBatchEffect model matrix with only an intercept term}
mat <- assay(vsd_remove_batch_intercept)
# only intercept term in model matrix, don't include Subpopulation or Tumor_JuxtaTumor
mm <- model.matrix(~1, colData(vsd_remove_batch_intercept))
# remove study batch effect
mat <- limma::removeBatchEffect(mat, batch = vsd_remove_batch_intercept$Study, design = mm)
assay(vsd_remove_batch_intercept) <- mat
pca_batch_correct_intercept <- plotPCA(vsd, intgroup = c("Subpopulation","Study")) + 
  aes(x = PC1, y = PC2, color = colData(vsd_remove_batch_intercept)$Subpopulation, shape =colData(vsd_remove_batch_intercept)$Study ) +
  labs(color='Subpopulation', shape = "Study") +
  ggtitle("PCA of VST data batch corrected using \nlimma::removeBatchEffect with intercept term only")
pca_batch_correct_intercept
```

```{r removeBatchEffect model matrix with only an intercept term, heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
# same template as previous heatmap
sampleDists <- dist(t(assay(vsd_remove_batch_intercept)))
sampleDistMatrix <- as.matrix( sampleDists )

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```

Limma's removeBatchEffect function requires a design matrix as input, this is the "treatment conditions" we wish to preserve. It is usually the design matrix with all experimental factors other than batch effects. The ideal scenario would be to include Subpopulation in this matrix. However, this treats `Unknown` as its own Subpopulation, and so will preserve differences between the InHouse samples and the other samples, as seen in the below PCA plot. This is contrary to what we want when assigning our samples to a cluster.

```{r removeBatchEffect tumor juxta + Subpopulation model matrix}
mat <- assay(vsd_remove_batch_tumor_juxta_subpopulation)
# create model matrix, full model matrix with Tumor_JuxtaTumor and Subpopulation
mm <- model.matrix(~Tumor_JuxtaTumor+Subpopulation, colData(vsd_remove_batch_tumor_juxta_subpopulation))
mat <- limma::removeBatchEffect(mat, batch = vsd_remove_batch_tumor_juxta_subpopulation$Study, design = mm)
assay(vsd_remove_batch_tumor_juxta_subpopulation) <- mat

pca_batch_correct_tumor_juxta_subpopulation <- plotPCA(vsd_remove_batch_tumor_juxta_subpopulation, intgroup = c("Subpopulation","Study")) + 
  aes(x = PC1, y = PC2, color = colData(vsd_remove_batch_tumor_juxta_subpopulation)$Subpopulation, 
      shape =colData(vsd_remove_batch_tumor_juxta_subpopulation)$Study ) +
  labs(color='Subpopulation', shape = "Study") +
  ggtitle("PCA of VST data batch corrected using \nlimma::removeBatchEffect with Tumor_JuxtaTumor\nand Subpopulation in model")
pca_batch_correct_tumor_juxta_subpopulation
```

```{r removeBatchEffect model matrix with Study+Subpopulation, heatmap}
# find euclidean distance between samples for heatmap generation (normalised data)
# same template as previous heatmap

sampleDists <- dist(t(assay(vsd_remove_batch_tumor_juxta_subpopulation)))
sampleDistMatrix <- as.matrix( sampleDists )

pheatmap(mat=sampleDistMatrix,
         show_rownames = FALSE,
         cluster_cols = TRUE,
         cluster_rows = TRUE,
         show_colnames = FALSE,
         annotation_col = subset_coldata,
         annotation_colors = ann_colors,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))

```

We can see from the PCA and heatmap above that including condition (tumour-juxtatumour) and Subpopulation in our model matrix leads to good separation of the Subpopulations, and a lack of clustering by batch. The problem with this approach is that it treats our `Unknown` Subpopulation as its own Subpopulation, which means it will cluster on its own no matter what we do. Is there a way around this? Frozen surrogate variable analysis could be an option, but this is usually used on microarray data, I don't know if it can be used on RNA-sequencing data.

```{r removeBatchEffect tumor juxta only matrix}
mat <- assay(vsd)
mm <- model.matrix(~Tumor_JuxtaTumor, colData(vsd))
mat_batch_removed_limma <- limma::removeBatchEffect(mat, batch = vsd$Study, design = mm)
```

In the PCA plot above, we have not told the `removeBatchEffect` function about our known Subpopulation, only whether the samples were taken from Tumor or Juxta-Tumor. It does not know to preserve differences between subpopulations when removing batch effects. Less variance being explained by PC1 than in the first scenario (39% vs 68%). In the first PCA plot, our in-house samples of unknown subpopulation cluster together on their own. 


```{r run BatchQC on our data}
#mat <- assay(vsd)
#modmatrix <- model.matrix(~as.factor(Tumor_JuxtaTumor), colData(vsd))
#batchQC(mat, batch=coldata$Study, condition= coldata$Subpopulation, 
      #  report_file = "batchqc_caf_data_not_corrected.html",
       # report_dir = ".", view_report = FALSE, interactive = FALSE
      #  )
```


```{r}
coldata_pca <- coldata
rownames(coldata_pca) <- coldata_pca$names
coldata_pca$names <- NULL
coldata_pca$files <- NULL
p <- pca(mat, metadata = coldata_pca)

```

```{r}
ggplot(p$rotated, aes(x = PC1, y = PC2, color = p$metadata$Study, shape = p$metadata$Subpopulation)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  labs(color='Study', shape = "Subpopulation") +
  ggtitle("PCA with transormed data after batch correction")
```

```{r Batch correction using Combat-Seq}
batch <- coldata$Study
mm <- model.matrix(~Tumor_JuxtaTumor, colData(vsd))
```


# Predicting subpopulation(s) present in In-House samples

## Assigning in-house samples to a CAF Subpopulation using K-nearest neighbours

1. Split mat into our known (training) and unknown (testing) subpopulations
2. Run KNN


```{r create separate variables for known and unknown CAF Subpopulations}
mat_t <- t(mat)
mat_known <- mat_t[coldata$names[which(coldata$Subpopulation != "Unknown")],]
coldata_known <- coldata[coldata$Subpopulation != "Unknown",]
mat_unknown <- mat_t[coldata$names[which(coldata$Subpopulation == "Unknown")],]
coldata_unknown <- coldata[coldata$Subpopulation == "Unknown",]
```

```{r}
library(class)
```

```{r}
##create a random number equal 90% of total number of rows
 ran <- sample(1:nrow(mat_known),0.9 * nrow(mat_known))
 ##training dataset extracted
 mat_train <- mat_known[ran,]
 
 ##test dataset extracted
 mat_test <- mat_known[-ran,]
```

```{r}
# these are our labels
caf_target_category <- coldata_known[ran,4]
caf_test_category <- coldata_known[-ran,4]
```

```{r run KNN}
pr <- knn(mat_train,mat_test,cl=caf_target_category,k=21)
```

```{r}
 tab <- table(pr,caf_test_category)
tab
```

```{r}
accuracy <- function(x){sum(diag(x)/(sum(rowSums(x)))) * 100}
 accuracy(tab)
```

```{r}
  outputs <- c()
for (i in 1:50){
  pr <- knn(mat_train,mat_test,cl=caf_target_category,k=i)
  number <- i
  tab <- table(pr,caf_test_category)
  accuracy_out <- accuracy(tab)
  outputs <- c(outputs, number = accuracy_out)
}
```

```{r}
plot(outputs)
```

```{r}
  prediction <- knn(mat_train,mat_unknown,cl=caf_target_category,k=13)
names(prediction) <- rownames(mat_unknown)
prediction
```

They are all predicted to be S1 using this initial application of the algorithm except one sample.
Possibly use interquartile range to improve performance?

## Deconvolution using CIBERSORTx

CIBERSORTx [@Newman2019] is the most commonly used tool for cell-type deconvolution. It is a machine learning method which carries out batch correction, correcting for between-platform differences in expression data. In this case, the signature matrix was made using all of the available CAF subpopulation data (40 S1 samples, 25 S3 samples, 24 S4 samples). It is possible to infer the proportions of the different subpopulations as well as a subpopulation-specific gene expression matrix for each sample. Either scRNA-seq, bulk RNA-seq or microarray data can be used as the reference.

- Create signature matrix for CIBERSORTx
- TPM normalise data,probably optional
- Run CIBERSORTx to figure out proportions of S1, S3 and S4 in In-house samples

The files for CIBERSORT (mixture file, reference data and phenotype classes file), were prepared using the `cibersortx_prepare_files.R` script.

`CIBERSORTx` was run using the following command:
```
docker run -v /home/kevin/Documents/PhD/cibersort/caf_subpopulation/infiles:/src/data -v /home/kevin/Documents/PhD/cibersort/caf_subpopulation/outfiles:/src/outdir cibersortx/fractions --username k.ryan45@nuigalway.ie --token b7f03b943ade9b4146dc2126b4ac9d19 --single_cell FALSE --refsample caf_subtypes_tpm_for_sig_matrix.txt --mixture caf_tpm_mixture.txt --rmbatchBmode TRUE --outdir /home/kevin/Documents/PhD/cibersort/caf_subpopulation/outfiles --phenoclasses /home/kevin/Documents/PhD/cibersort/caf_subpopulation/infiles/phenoclasses_caf.txt
```

```{r CIBERSORT initial results}
cibersort_results <- read.table("/home/kevin/Documents/PhD/cibersort/caf_subpopulation/outfiles/CIBERSORTx_Adjusted.txt", header = T)
cibersort_results
```

Results look strange with P-value of 9999. CIBERSORTx was run using the Docker image and using the GUI, and different results were obtained. When the Docker image was used and the number of permutations was changed from 0 to 100, the p-value changed to 0.000. The proportions look different between the two methods, with the GUI predicting about 0.75 S1 with the rest being S4 for most samples.


## References
