---
title: "Plot genes of interest CAF subpopulations"
author: "Kevin Ryan"
date: "2023-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DESeq2)
library(ggplot2)
library(biomaRt)
```

## Intro

Goal: plot expression of 5 genes of interest using DESeq2's plotCounts() function
Do for both batch-corrected and not corrected data.
Indicate whether they were differentially expressed.

```{r}
mart <- useMart(biomart = "ensembl", dataset = "hsapiens_gene_ensembl", #host="https://www.ensembl.org")
                host="uswest.ensembl.org")

ensembl.version.2.hgnc <- getBM(attributes = c("ensembl_gene_id_version", "ensembl_gene_id", "hgnc_symbol", "entrezgene_id"), mart = mart, useCache = FALSE)
genes.interest <- c("WT1", "SLC7A2", "GPR37", "ADGRF5", "PCDH10")
gene.interest.ensg.version <- filter(ensembl.version.2.hgnc, hgnc_symbol %in% genes.interest)
gene.interest.ensg.version
```

### Read in data

### batch-corrected data

```{r Read in batch corrected data ENSG and remove inhouse data as subpopulation is unknown}
dds.ensg.remove.outliers.batch.corrected <- readRDS("../intermediate_files/dds_batch_corrected_group_tumor_ensembl_gene_id_version_2022-10-07.Rds")
keep <- which(colData(dds.ensg.remove.outliers.batch.corrected)$Subpopulation != "Unknown")
dds.ensg.remove.outliers.batch.corrected.no.inhouse <- dds.ensg.remove.outliers.batch.corrected[,keep]
colData(dds.ensg.remove.outliers.batch.corrected.no.inhouse)$Tumor_JuxtaTumor <- as.factor(colData(dds.ensg.remove.outliers.batch.corrected.no.inhouse)$Tumor_JuxtaTumor)
colData(dds.ensg.remove.outliers.batch.corrected.no.inhouse)$Subpopulation <- as.factor(colData(dds.ensg.remove.outliers.batch.corrected.no.inhouse)$Subpopulation)
dds.ensg.remove.outliers.batch.corrected.no.inhouse <- DESeq(dds.ensg.remove.outliers.batch.corrected.no.inhouse)
dds.ensg.remove.outliers.batch.corrected.no.inhouse <- filter_out_low_expressed(dds.ensg.remove.outliers.batch.corrected.no.inhouse)
```

### non batch-corrected data

```{r}
dds.ensg.remove.outliers.no.inhouse <- readRDS("../intermediate_files/dds_not_corrected_remove_outliers_ensg_2022-10-26.Rds")
```

### DE analysis results

differential expression analysis using non batch-corrected data and 9 surrogate variables (correlation subpopulation sig, correlation batch not sig.)

```{r}
dds.ensg.remove.outliers.no.inhouse.full.design <- readRDS("../intermediate_files/dds_model_fit_ensg_nsv9_2022-10-26.Rds")
```


## Make plots

```{r}
genes.interest <- gene.interest.ensg.version$hgnc_symbol
genes.interest.ensg <- gene.interest.ensg.version$ensembl_gene_id_version

# genes_interest_common_names <- c("FAP", "CD29", "αSMA", "PDPN", "PDGFRB", "FSP1", "CAV1")
dds_batch_corrected_outliers_removed_marker_genes <- dds.ensg.remove.outliers.no.inhouse[genes.interest.ensg,]
plotcounts_dfs <- list()
for (i in 1:length(genes_interest)){
  plotcounts_dfs[[i]] <- plotCounts(dds_batch_corrected_outliers_removed_marker_genes, 
                                    gene=genes_interest[i], 
                                    returnData = TRUE,
                                    intgroup = c("Subpopulation", "Tumor_JuxtaTumor")
                                    )
}

plots_out_tumor_juxtatumor <- list()
for (i in 1:length(plotcounts_dfs)){
  plt <- caf_plot_tumour_juxtatumour(df_for_plotting = plotcounts_dfs[[i]], 
                                                                   gene = genes_interest_common_names[i])
  plots_out_tumor_juxtatumor[[i]] = plt
}
```

```{r}
ggar_obj_tumor_juxtatumor <- ggarrange(plotlist = plots_out_tumor_juxtatumor, common.legend = TRUE) # rel_heights values control title margins
ggar_obj_tumor_juxtatumor_annotated <- annotate_figure(ggar_obj_tumor_juxtatumor, bottom = text_grob("CAF Subpopulation"))
ggar_obj_tumor_juxtatumor_annotated
```
